---
title: "Genetic Recombination"
author: "Chris Gaynor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genetic Recombination}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(AlphaSimR)
```

This article provides an overview of the biology of genetic
recombination and how it is modeled within AlphaSimR. I will first
provide a brief review of meiosis with special emphasis on how the
process differs between diploid and autopolyploid species. I will then
discuss genetic maps before concluding with an overview of the specific
models used to simulate genetic recombination and of the machinery
AlphaSimR uses to track the results.

Recombination is the single most important process in a breeding
simulation. Selection acts on individuals, but progress is only possible
because meiosis reassembles parental haplotypes into new combinations.
Everything AlphaSimR does downstream of a cross - linkage disequilibrium,
the Bulmer effect, the decay of segregating variance, the accuracy of
genomic prediction across generations - depends on getting this one
process right.

## Meiosis

Meiosis is the cell division that produces gametes. In a diploid, each
somatic cell carries two copies of every chromosome, one inherited from
the mother and one from the father. Gametes carry only one copy, so
meiosis must halve the chromosome number while producing cells that are
genetically distinct from one another and from the parent.

The process begins with DNA replication. Each chromosome is duplicated
to produce two identical **sister chromatids** joined at the centromere.
A diploid cell entering meiosis therefore contains four chromatids for
each chromosome: two from the maternal chromosome and two from the
paternal chromosome.

During prophase of the first meiotic division, homologous chromosomes
find each other and pair along their length. The paired structure is
called a **bivalent**, and because it contains four chromatids it is also
referred to as a **tetrad**. Physical connections form between
non-sister chromatids at points called **chiasmata**. A chiasma is the
cytological signature of a **crossover**: a reciprocal exchange of
material between the two chromatids involved.

Two features of this arrangement matter a great deal for simulation.
First, a chiasma is an event of the four-strand bundle, not of an
individual chromatid. Each chiasma involves exactly two of the four
chromatids, so any particular chromatid has a one half chance of taking
part in any particular chiasma. Second, in the absence of *chromatid
interference*, the pair of chromatids involved in each chiasma is chosen
independently of the pairs involved in every other chiasma. AlphaSimR
assumes no chromatid interference, which is well supported empirically
and which makes the second step above a simple independent coin flip per
chiasma.

The first meiotic division separates the homologous chromosomes. This is
the *reductional* division and it is where independent assortment
occurs: the orientation of each bivalent on the spindle is independent of
every other bivalent. The second meiotic division separates the sister
chromatids, the *equational* division, and yields four haploid cells,
each containing a single chromatid of each chromosome.

The consequence for a simulator is a convenient separation of concerns. A
gamete is a mosaic of the two parental chromosomes, and the positions at
which the mosaic switches from one parent to the other are the crossovers
that happened to involve the chromatid that ended up in that gamete.
Simulating a gamete therefore requires only two things: a rule for where
chiasmata fall along the chromosome, and a rule for which chromatids each
chiasma involves.

### Meiosis in polyploids

There are two primary types of polyploids: autopolyploids and
allopolyploids.

Allopolyploids, such as bread wheat, carry multiple genome copies derived
from different ancestral species. The copies are similar enough to
coexist in the same nucleus but different enough that each chromosome has
only one true homologue. Pairing is therefore *disomic*: chromosomes form
ordinary bivalents and meiosis proceeds essentially as it does in a
diploid. From a modeling standpoint an allopolyploid can be treated as a
collection of independent diploid genomes, and AlphaSimR handles it that
way.

Autopolyploids, such as cultivated potato or alfalfa, carry multiple
copies of the *same* genome. Every chromosome has more than one equally
valid pairing partner, and this changes meiosis in ways that have no
diploid analogue. In an autotetraploid, the four homologues of a
chromosome may pair as two independent bivalents, or all four may
associate into a single **quadrivalent**. Which of these happens is
partly a property of the species and partly of the chromosome, and it is
usually summarised as a probability of quadrivalent formation.

Quadrivalent pairing has two consequences. It allows a gamete to inherit
material from more than two of the four homologues, and it makes **double
reduction** possible. Double reduction occurs when a gamete receives two
copies of the same parental chromosome segment, that is, when the two
chromosomes a gamete inherits are sister chromatids over some interval.
This cannot happen under bivalent pairing. Under quadrivalent pairing it
happens for loci that are separated from the centromere by a crossover,
so its frequency increases with distance from the centromere. Because the
centromere position determines the pattern, AlphaSimR requires centromere
positions for autopolyploid simulations, which is why the `MapPop` class
carries a `centromere` slot.

## Genetic maps

The raw material for a recombination model is a genetic map. The idea
goes back to Morgan's fly lab, where Sturtevant recognised that the
frequency with which two markers are separated during meiosis could be
used to order them along a chromosome and to assign them positions on a
linear scale. That scale is measured in Morgans; one Morgan is the
distance over which one crossover is expected per gamete per meiosis.

A genetic map is not the same as a physical map. Physical distance is
measured in base pairs and reflects the actual DNA sequence. Genetic
distance reflects how often recombination separates loci, and the
relationship between the two varies enormously along a chromosome,
typically suppressed near centromeres and elevated near telomeres.
AlphaSimR works exclusively in genetic distance, and every position it
stores is in Morgans.

The quantity that is actually observed in a mapping experiment is the
**recombination fraction** $r$, the probability that two loci are
separated in a gamete. Map distance $d$ and recombination fraction are
not the same thing, because an even number of crossovers between two loci
leaves them in the parental configuration and is invisible. A **mapping
function** converts between the two. If crossovers occur independently as
a Poisson process, then

$$r = \frac{1 - e^{-2d}}{2}$$

which is Haldane's mapping function. Haldane's function assumes no
crossover interference, and real chromosomes show interference: the
occurrence of one crossover suppresses others nearby. Kosambi's mapping
function

$$r = \frac{1}{2}\tanh(2d)$$

builds in a moderate amount of interference and generally fits observed
data better. These functions matter here because they describe the
behaviour a recombination model should reproduce, and because AlphaSimR's
default parameterisation is chosen to approximate Kosambi.

In AlphaSimR a genetic map is a list with one element per chromosome.
Each element is a named numeric vector giving the position of every
segregating site in Morgans, with the first site placed at zero. The map
is held by the `SimParam` object and can be inspected with `getGenMap`.

```{r}
founderPop = quickHaplo(nInd = 10, nChr = 2, segSites = 5, genLen = 1)
SP = SimParam$new(founderPop)

head(getGenMap(SP))
```

The map used for meiosis can be replaced with `SP$switchGenMap`, and
male and female maps can be set separately with `SP$switchFemaleMap` and
`SP$switchMaleMap`. Sex-specific recombination rates are common in
animals, and where the recombination *landscape* is assumed to be the
same in both sexes but its total length differs, `SP$setRecombRatio`
provides a shortcut that rescales both maps and both centromere positions
around the current map.

```{r}
SP$setRecombRatio(2) # Twice as much recombination in females

sapply(SP$femaleMap, max)
sapply(SP$maleMap, max)
```

## Recombination models

Everything in this section concerns a single chromosome of total genetic
length $L$ Morgans. AlphaSimR simulates each chromosome independently,
which is exactly the independent assortment described earlier, and it
does so in parallel when OpenMP is available. Each chromosome is given
its own random number substream so that results are reproducible
regardless of how many threads are used.

Simulating a gamete proceeds in three steps. Chiasmata are placed along
the chromosome; each chiasma is thinned with probability one half to
decide whether the focal chromatid took part in it; and the surviving
crossover positions are converted into a set of intervals that are copied
from one parental haplotype or the other.

The thinning step is what connects the chiasma process to the crossover
process. Because a gamete sees only half the chiasmata, a chiasma rate of
two per Morgan is required to give the one crossover per Morgan that
defines the Morgan. Every model below therefore places chiasmata at rate
$2$ per Morgan; the difference between them is entirely in how those
chiasmata are spaced.

Three models are available, and all three are controlled through two
`SimParam` fields: `v`, the interference parameter, and `p`, the
proportion of crossovers arising from a non-interfering pathway.

### The count-location model

The simplest model treats chiasmata as a homogeneous Poisson process. The
number of chiasmata on the chromosome is drawn from a Poisson
distribution with mean $2L$, and each is placed at a position drawn
uniformly along the map. There is no interference: the position of one
chiasma tells you nothing about the position of any other. The resulting
crossover process is exactly the one underlying Haldane's mapping
function.

This model is used when `p` is set to 1, meaning all crossovers come from
the non-interfering pathway. It is the cheapest of the three and it is
the right choice when the simulation is not sensitive to interference or
when a Haldane map is what is wanted.

### The gamma model

Interference is easier to model in terms of the *spacing* between
chiasmata than in terms of their positions. The gamma model, developed
for this purpose by Broman, Weber, Zhao and others, treats chiasmata
along a chromosome as a stationary renewal process whose inter-arrival
distances follow a gamma distribution with shape $v$ and rate $2v$. The
mean spacing is then $1/2$ Morgan, giving the required chiasma rate of two
per Morgan, and the shape parameter alone controls interference.

When $v = 1$ the gamma distribution is exponential, the renewal process
is Poisson, and the model reduces to the count-location model with no
interference. As $v$ increases the spacing distribution concentrates
around its mean, short gaps become rare, and interference strengthens.
The AlphaSimR default is $v = 2.6$, which is the value that makes the
resulting map function closely approximate Kosambi's.

```{r}
SP$v # Interference parameter, 2.6 approximates Kosambi
SP$p # Proportion of non-interfering crossovers
```

One implementation detail is worth mentioning because it is a common
source of bias in renewal-process simulations. A renewal process started
at position zero is not stationary: the first interval is an ordinary
gamma deviate, whereas the interval that happens to straddle an arbitrary
point on a long chromosome is length-biased and stochastically larger.
The usual remedy is to burn in from several Morgans upstream and discard
everything before position zero, which costs twenty or so wasted random
deviates per chromosome. AlphaSimR instead samples the equilibrium
directly. The interval covering position zero is drawn from a
$\mathrm{Gamma}(v + 1, \text{scale})$ distribution, which is exactly the
length-biased version of the spacing distribution, and position zero is
placed uniformly within it. Splitting that interval at the uniform point
gives the forward and backward recurrence times directly, using two
deviates instead of twenty and with no approximation.

### The gamma sprinkling model

There is good evidence in several species that crossovers arise from two
distinct pathways. The dominant pathway is subject to interference; a
second, minor pathway is not. The gamma sprinkling model, sometimes
called the two-pathway model, accommodates this by combining the two
processes.

A proportion $p$ of crossovers is assigned to the non-interfering
pathway. Interfering chiasmata are drawn from a gamma renewal process
whose spacings have shape $v$ and gamma rate parameter $2v(1-p)$. That
gives a mean spacing of $1/(2(1-p))$ Morgans, and so an arrival rate of
$2(1-p)$ chiasmata per Morgan. Non-interfering chiasmata arrive as a
Poisson process at $2p$ per Morgan; because a Poisson process is
memoryless it needs no stationary start. The two sets of positions are
merged and sorted, and the total rate is again two per Morgan.

Setting $p = 0$ recovers the pure gamma model, and setting $p = 1$
recovers the count-location model, so the three models form a single
family. To use the sprinkling model, set `p` to a value between zero and
one.

```{r}
SP$p = 0.1 # 10% of crossovers from a non-interfering pathway
```

### From crossovers to haplotypes

Once crossover positions are known, the gamete is assembled. AlphaSimR
searches the genetic map to find which marker interval each crossover
falls into, and records the resulting mosaic as a two-column matrix. The
first column identifies the parental haplotype the segment is copied
from, and the second gives the segregating site at which that segment
begins. Genotypes are stored as packed bits, and segments are copied
between haplotypes in whole bytes wherever possible, which is why a
crossover-heavy simulation is not much slower than a crossover-free one.

A subtlety arises here. Crossovers are sampled on the continuous genetic
map, but the simulation only stores genotypes at segregating sites. Two
crossovers falling within the same marker interval leave no trace in the
data; neither does a pair of crossovers that returns the gamete to the
chromosome it was already reading from. These *hidden* crossovers are
removed from the recorded history, so
what AlphaSimR reports is the set of crossovers that are actually
observable given the density of the map. This keeps the recorded history
consistent with the genotypes and keeps its size proportional to marker
density rather than to map length.

### Recombination models in autopolyploids

For an autopolyploid, meiosis is simulated one pairing group at a time.
The homologues of a chromosome are shuffled and then processed in groups
of four. Each group of four either forms two independent bivalents or a
single quadrivalent, with the quadrivalent chosen with probability
`SP$quadProb`. If fewer than four homologues remain, the remainder is
handled as a bivalent. An autotetraploid therefore has one group of four;
an autohexaploid has one group of four and one bivalent.

```{r}
SP$quadProb # Probability of quadrivalent pairing, default is no quadrivalents
```

Bivalents within a polyploid are simulated exactly as they are in a
diploid, using whichever of the three models `v` and `p` specify. The
quadrivalent needs its own treatment.

AlphaSimR models the quadrivalent as a *cross-type* configuration. An
exchange point is sampled uniformly along the chromosome, and on either
side of it the four homologues pair up in a different arrangement. This
gives four pairing arms: the heads of chromosomes 1 and 2, the tails of
chromosomes 2 and 3, the heads of chromosomes 3 and 4, and the tails of
chromosomes 1 and 4. Chiasmata are sampled along all four arms with
interference operating *between* arms as well as within them, since the
four arms are part of a single synapsed structure. One arm is chosen at
random and sampled from the stationary process described above; the
remaining arms are then sampled conditional on the distance to the
nearest chiasma already placed, so that the interference constraint
carries across the whole configuration. When `p` is non-zero,
non-interfering chiasmata are sprinkled onto each arm afterwards.

Each chiasma is assigned a pair of chromatids independently, as in the
bivalent case. A quadrivalent yields two of the gamete's chromosomes, and
these are determined by choosing two centromeres. AlphaSimR always takes
the centromere of the first chromosome and its first chromatid, and
samples the second centromere from the remaining three chromosomes and
either of their chromatids. Each gamete is then resolved by working
backwards from its centromere to the head of the chromosome to establish
which chromatid it starts on, and then forwards from head to tail
recording every chiasma that involves the chromatid currently being
followed.

Because the two retained centromeres always come from different
homologues, the two chromosomes a gamete receives are never sister
chromatids at the centromere itself. Distal to a crossover, however, they
may both trace back to the same parental chromosome, and this is double
reduction. Its frequency is zero at the centromere and increases with
genetic distance from it, which is the pattern expected from theory and
observed in practice.

## Recombination tracking in AlphaSimR

Everything described above is recorded if you ask for it. Recombination
tracking is off by default because the records accumulate over
generations; turn it on before creating any populations.

```{r}
set.seed(5678)

founderPop = quickHaplo(nInd = 10, nChr = 1, segSites = 20, genLen = 1)

SP = SimParam$new(founderPop)
SP$setTrackRec(TRUE)

basePop = newPop(founderPop, simParam = SP)
pop = randCross(basePop, nCrosses = 4, simParam = SP)
```

Recombination tracking implies pedigree tracking, and `SP$setTrackRec`
turns pedigree tracking on automatically. Turning pedigree tracking off
with `SP$setTrackPed(FALSE)` turns recombination tracking off with it,
since a recombination history is meaningless without the parents it
refers to.

The raw records are held in `SP$recHist`, a list indexed by individual.
Each element is a list over chromosomes, and each of those a list over
the individual's chromosome copies. The innermost element is the
two-column matrix described earlier.

```{r}
SP$recHist[[pop@id[1]]][[1]][[1]] # First individual, chromosome 1, first copy
```

The first column gives the parental haplotype and the second the
segregating site at which that haplotype starts contributing. A matrix
with a single row means the gamete was inherited without any observable
crossover. Note that the parental haplotype numbering is relative to the
parent that contributed the gamete, so a `1` in the first copy refers to
the mother's first haplotype and a `1` in the second copy refers to the
father's first haplotype.

The raw histories describe one meiosis. More often what is wanted is the
origin of each segment in terms of the founders, several generations
back. `pullIbdHaplo` provides this by walking the pedigree, composing the
recombination histories along the way, and returning a matrix in which
every locus is labelled with the founder haplotype it descends from.

```{r}
pullIbdHaplo(pop, simParam = SP)
```

Each row is one haplotype of one individual, named `id_copy`, and each
column is a segregating site. Runs of identical values are
identical-by-descent segments, and the points at which the values change
are the accumulated crossovers separating this individual from the
founders. Founder haplotypes are numbered in order of creation, so a base
population of ten diploid individuals occupies labels 1 through 20.

This is the natural way to see recombination in a simulation. The
haplotypes themselves often hide it: a crossover between two markers that
happen to carry the same allele leaves no visible trace in the genotype
data, and with realistic marker densities most crossovers fall between
markers that are not informative. The IBD representation shows the
crossover regardless.

The same tracking machinery is used by every function that performs a
meiosis, not just crossing. Doubled haploid production with `makeDH`
simulates a single gamete and then doubles it, so both haplotypes of a DH
line share one recombination history. `reduceGenome` samples gametes
without combining them, and `mergeGenome` and `doubleGenome` manipulate
ploidy directly. All of them contribute to `SP$recHist` and are visible
through `pullIbdHaplo`.

```{r}
dh = makeDH(pop, nDH = 1, simParam = SP)

pullIbdHaplo(dh, simParam = SP)
```

Both copies of each DH line are identical, as expected. Each is built
from the two haplotypes of the parent it came from, and where the gamete
happened to recombine both copies carry the same crossovers. A gamete
that recombined nowhere gives a DH line that simply doubles one parental
haplotype.

## Summary

AlphaSimR simulates recombination by placing chiasmata along a genetic
map at a rate of two per Morgan, thinning them by one half to obtain the
crossovers seen by a single chromatid, and assembling gametes as mosaics
of the parental haplotypes. Interference is controlled by a single shape
parameter `v`, with the default of 2.6 approximating a Kosambi map, and a
second parameter `p` adds a non-interfering pathway. Setting `p` to 1
removes interference entirely and reduces the model to the
count-location model.

Autopolyploids are handled by pairing homologues in groups of four, each
of which forms either two bivalents or a quadrivalent with probability
`SP$quadProb`. The quadrivalent uses a cross-type configuration with
interference operating across all four arms, and produces double
reduction at a rate that increases with distance from the centromere.

Turning on `SP$setTrackRec(TRUE)` records every observable crossover, and
`pullIbdHaplo` converts those records into founder-haplotype origins for
any individual in the simulation.
