---
title: "Genomic Selection in AlphaSimR"
author: "Chris Gaynor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: AlphaSimR.bib
vignette: >
  %\VignetteIndexEntry{Genomic Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(AlphaSimR)
```

## Introduction

Genomic selection is selection on a prediction of an individual's genetic
merit made from its marker genotypes. AlphaSimR provides a family of
ridge regression BLUP (RR-BLUP) models for making these predictions
[@AlphaSim; @gaynor_2017]. They are fitted in C++ and are meant to be
re-fitted many times over the course of a simulation, so they are
deliberately narrow in scope: they are fast, they take their data
directly from a population object, and they return an object that
AlphaSimR knows how to turn back into estimated values for any
population you like.

This article is mostly about *using* those functions: which one to call,
what to hand it, and where the call belongs in a breeding cycle. The
model each one fits is described later, in a section you can skip on a
first reading.

Two earlier articles are assumed. The article on the genome explains how
QTL and SNP chips are defined and how genotypes are stored and pulled.
The article on quantitative genetics explains the distinction between a
genetic value and a breeding value, and why a breeding value is a
property of a *reference population* rather than of an individual on its
own. That distinction reappears below, because it determines what
`setEBV` gives you.

## Setting up an example

The example below is used throughout.

```{r}
set.seed(1234)

founderPop = runMacs(nInd = 500, nChr = 10, segSites = 100)

SP = SimParam$new(founderPop)
```
```{r include=FALSE}
SP$nThreads = 1L
```
```{r}
# Keep the QTL and the SNP chip on separate segregating sites
SP$restrSegSites(minQtlPerChr = 20, minSnpPerChr = 50, overlap = FALSE)

SP$addTraitAD(nQtlPerChr = 20, meanDD = 0.4)
SP$addSnpChip(nSnpPerChr = 50)
SP$setVarE(h2 = 0.4)

pop = newPop(founderPop, simParam = SP)
```

Two choices here are worth copying into your own simulations.

`runMacs` rather than `quickHaplo`. Genomic prediction works because
markers are in linkage disequilibrium with QTL, and `quickHaplo` samples
haplotypes at linkage equilibrium. A simulation built on `quickHaplo`
will show almost no predictive ability unless the QTL themselves are
used as markers, which makes it a poor choice for demonstrating or
benchmarking genomic selection.

`restrSegSites` to force the QTL and the SNP chip onto different
segregating sites. Without it the two sets are sampled independently and
may overlap, which quietly inflates accuracy.

## The three steps

Every genomic selection analysis in AlphaSimR has the same shape.

1. Assemble a **training population** carrying both phenotypes and
   genotypes.
2. **Fit a model** to it, producing an object of `RRsol-class`.
3. **Apply that solution** to the selection candidates with `setEBV`,
   which fills their `ebv` slot.

Selection then proceeds with the ordinary selection functions, using
`use="ebv"`.

```{r}
# 1. Training population: phenotyped and genotyped
trainPop = setPheno(pop[1:400], simParam = SP)

# Selection candidates: genotyped only
candidates = pop[401:500]

# 2. Fit
ans = RRBLUP(trainPop, simParam = SP)

# 3. Predict
candidates = setEBV(candidates, ans, simParam = SP)

# Accuracy, measured on individuals that did not train the model
cor(ebv(candidates)[, 1], gv(candidates)[, 1])
```

A training population is an ordinary `Pop-class` object. There is no
special class and no registration step; a model reads the response and
the genotypes out of whatever population you hand it. The only
requirements are that the population carries a value to train on and
that the SNP chip you intend to use was defined before the population
was created.

The response is chosen with the `use` argument, which accepts `"pheno"`
(the default), `"gv"`, `"bv"`, `"ebv"` or `"rand"`. Training on
phenotypes is the realistic choice. Training on `"gv"` or `"bv"` is a
diagnostic: it removes environmental noise and shows what the model
could achieve if phenotyping were perfect. `"rand"` trains on noise and
is useful for confirming that an apparent gain really comes from the
prediction.

Training populations are usually assembled from several sources, and the
natural tool for that is `c()`:

```{r}
combined = c(setPheno(pop[1:200], reps = 1, simParam = SP),
             setPheno(pop[201:400], reps = 3, simParam = SP))

nInd(combined)
```

## Choosing a model

The first question is not which algorithm to use but **what you are
trying to predict**. That answer picks the model; everything else is a
matter of memory and speed.

| You want to predict | Use | Ask `setEBV` for |
|---|---|---|
| Merit of a parent, additive trait | `RRBLUP` | `"bv"` or `"gv"` |
| Several correlated traits at once | `RRBLUP` with `traits=1:n` | `"bv"` or `"gv"` |
| Performance of a line or clone, with dominance | `RRBLUP_D` | `"gv"` |
| GCA of inbreds in a two-pool hybrid program | `RRBLUP_GCA` | `"female"`, `"male"` |
| Performance of a specific single cross | `RRBLUP_SCA` | `"gv"` |
| Any of the above, training population too large to fit | `fastRRBLUP` | as above |

**`RRBLUP`** is the default choice. It fits additive marker effects and
is the only model in the family that fits several traits at once;
supplying more than one trait fits a multivariate model, which borrows
information across correlated traits at the cost of estimating a full
covariance matrix. If you are unsure, start here.

**`RRBLUP_D`** adds a dominance covariate for each marker. Reach for it
when the trait was defined with dominance *and* you care about the
genetic value rather than only the breeding value — predicting the
performance of a cross or a clone, for instance, rather than the merit
of a parent [@xiang_2016]. If you only ever select parents on additive
merit, the extra term buys you little.

**`RRBLUP_GCA`** and **`RRBLUP_SCA`** are for hybrid programs with two
heterotic pools. Rather than one set of additive effects they estimate a
set for each pool, so an allele can be worth something different
depending on which parent contributed it. `RRBLUP_SCA` adds a dominance
term on top, so specific combining ability can be predicted as well as
general combining ability [@Massman13]. Both fill the `female` and
`male` slots and leave `bv` empty. In practice they are paired with
`setPhenoGCA`, which phenotypes candidate parents through testcrosses,
and with `hybridCross` for producing the crosses themselves.

@dejong_2023 compared these models for selecting inbred parents on
predicted GCA in a simulated maize program, and found that models
fitting pool-specific effects together with dominance gave the best fit
and the highest genetic gain. They also found that the ranking depends
on marker density, with the models differing more in genetic gain at low
density and more in heterosis at high density — a useful warning against
assuming one model is uniformly best. The help page for `RRBLUP_SCA`
notes that no consistent benefit over `RRBLUP_GCA` has been observed in
AlphaSimR's own testing, so it is worth checking both in your own
simulation rather than assuming the more complex model wins.

**`fastRRBLUP`** is the escape hatch for size. It solves the same
additive model as `RRBLUP` but never forms a square matrix, so its
memory cost stays close to the cost of the genotypes themselves. Use it
when the training population has grown large enough that the others will
not fit. It is the only member of the family whose cost does not blow up
with the number of records.

### Fitting the same model more cheaply

Every model above has a numbered counterpart: `RRBLUP2`, `RRBLUP_D2`,
`RRBLUP_GCA2`, `RRBLUP_SCA2`. These fit *exactly the same models* and
differ only in how the equations are solved. The practical reason to
care is that the numbered versions let you supply the variance
components and skip estimating them:

```{r}
ans2 = RRBLUP2(trainPop, Vu = ans@Vu[1, 1], Ve = ans@Ve[1, 1],
               useEM = FALSE, simParam = SP)

candidates = setEBV(candidates, ans2, simParam = SP)

cor(ebv(candidates)[, 1], gv(candidates)[, 1])
```

With `useEM=FALSE` and `Vu` and `Ve` supplied, the work reduces to a
single solve of the mixed model equations, which is quick. The common
pattern is to fit `RRBLUP` once at the start of a simulation, keep its
`Vu` and `Ve`, and use `RRBLUP2` with those values for every subsequent
re-training. Whether holding the variance components fixed is acceptable
depends on how far the population drifts over the simulation, and that
is a question only the simulation can answer — check against a full
re-fit before relying on it.

If you let the numbered models estimate the variance components
themselves they use an EM algorithm, which is slower and less reliable
than the approach the unnumbered models use. `maxIter` defaults to a
value that is usually too small to converge, and a warning is issued
when it does not. Left to estimate their own variance components, the
numbered models are rarely the better choice.

## Using the models in a breeding program

The three steps above are the whole API. What takes judgement is where
they go in a breeding cycle and what feeds them. @bancic_2025 work
through complete plant breeding programs in AlphaSimR — line, clonal and
hybrid — each with a genomic selection variant, and the patterns below
follow that paper. Its scripts are a good companion to this article.

The examples in this section are sketches rather than runnable code;
they assume a pipeline of yield trial stages with the usual names.

### Replacing a phenotyping stage

The most direct use of genomic selection is to predict individuals that
have not been phenotyped, which lets an early phenotyping stage be
dropped altogether. In a wheat line program the headrow stage exists
only to screen doubled haploids, so predicting the doubled haploids
removes the need for it:

```{r eval=FALSE}
# Predict every DH and advance the best to the first yield trial
DH = setEBV(DH, gsModel, simParam = SP)

PYT = selectWithinFam(DH, famMax, use = "ebv", simParam = SP)
PYT = selectInd(PYT, nPYT, use = "ebv", simParam = SP)
PYT = setPheno(PYT, varE = varE, reps = repPYT, simParam = SP)
```

Note the two-stage selection. Selecting within families first, then
across them, limits how many entries any one cross can contribute.
Genomic predictions are strongly correlated within a family, so
unrestricted truncation selection on EBVs tends to fill the next stage
with full sibs and burn through genetic variance faster than the
equivalent phenotypic scheme.

### Selecting parents earlier

The larger gain usually comes from shortening the cycle rather than from
predicting more accurately. Once doubled haploids can be ranked without
phenotyping, parents can be drawn from them instead of from the elite
trial several years later:

```{r eval=FALSE}
DH = setEBV(DH, gsModel, simParam = SP)

newParents = selectInd(DH, nNewParents, use = "ebv", simParam = SP)

# Replace the oldest parents with the newly predicted ones
Parents = c(Parents[-(1:nNewParents)], newParents)
```

Taken further, this becomes the two-part strategy of @gaynor_2017: a
product development component that still runs on phenotypes, and a
population improvement component that runs entirely on EBVs and cycles
several times per year.

### Maintaining a training population

A model is only as good as what trained it, and in a running simulation
the training population has to be maintained rather than defined once.
The usual arrangement accumulates records during a burn-in and then
holds the size constant with a sliding window, dropping the oldest
records as new ones arrive:

```{r eval=FALSE}
if (year == startTP) {
  trainPop = c(PYT, AYT, EYT)
} else if (year <= nBurnin) {
  # Accumulate
  trainPop = c(trainPop, PYT, AYT, EYT)
} else {
  # Fixed size: drop the oldest year of records
  nNew = c(PYT, AYT, EYT)@nInd
  trainPop = c(trainPop[-(1:nNew)], PYT, AYT, EYT)
}
```

Capping the size matters for more than run time. Records from many
generations back describe a population that no longer exists, and the
linkage disequilibrium they carry has been broken up by recombination
since. A window of the last few years is both cheaper and, often, more
accurate than everything ever recorded.

### How often to retrain

Refitting every cycle on an updated training population is the
straightforward choice and the one the examples above assume. It is also
the expensive one. Three cheaper options, in rough order of how much
they give up:

- Refit every cycle with `RRBLUP2` and fixed `Vu` and `Ve`, taken from
  one full `RRBLUP` fit at the start.
- Refit every cycle with `fastRRBLUP`, whose cost grows far more slowly
  with the number of records.
- Refit every *k* cycles and reuse the solution in between. A solution
  is an ordinary object; nothing stops you applying it to several
  generations of candidates.

The following runs three cycles of genomic selection on the example
population, refitting each cycle against a training population that
accumulates the newest phenotyped generation:

```{r}
gsPop = pop
gsTrainPop = setPheno(pop, simParam = SP)

for (cycle in 1:3) {
  # Fit to everything phenotyped so far
  gsModel = RRBLUP(gsTrainPop, simParam = SP)

  # Predict and select
  gsPop = setEBV(gsPop, gsModel, simParam = SP)
  gsPop = selectCross(gsPop, nInd = 50, nCrosses = 100, nProgeny = 2,
                      use = "ebv", simParam = SP)

  # Phenotype the new generation and add it to the training records
  gsTrainPop = c(gsTrainPop, setPheno(gsPop, simParam = SP))

  cat("cycle", cycle, " meanG =", round(meanG(gsPop), 3),
      " nTrain =", nInd(gsTrainPop), "\n")
}
```

The mean genetic value rises each cycle, and the training population
grows as it does — which is exactly the cost that the options above are
there to control.

### Hybrid programs

In a two-pool hybrid program the training records are testcross
performance rather than line performance, and the predictions wanted are
combining abilities. `RRBLUP_GCA` fits both pools at once and `setEBV`
is then asked for the pool-specific value:

```{r eval=FALSE}
gsModel = RRBLUP_GCA(hybridTrainPop, simParam = SP)

maleDH = setEBV(maleDH, gsModel, value = "male", simParam = SP)
femaleDH = setEBV(femaleDH, gsModel, value = "female", simParam = SP)

newMaleParents = selectInd(maleDH, 10, use = "ebv", simParam = SP)
newFemaleParents = selectInd(femaleDH, 10, use = "ebv", simParam = SP)
```

A simpler alternative, which @bancic_2025 use as their default, is to
fit an ordinary `RRBLUP` to the hybrid training population and read the
result as a general combining ability. That works because a
hybrid-trained additive model estimates the average effect of an allele
across the opposite pool, which is what GCA is. Fitting `RRBLUP`
separately within each pool on inbred performance is a third option, and
is the one to use when testcross data are not available.

To predict the performance of a *specific* cross rather than the merit
of its parents, fit `RRBLUP_SCA` and ask for `value="gv"` on the hybrid
population produced by `hybridCross`.

## Working with EBVs

`setEBV` applies a solution to a population and writes the result into
the `ebv` slot. The `value` argument chooses what is estimated: `"gv"`
(the default) for an estimated genetic value, `"bv"` for an estimated
breeding value, and `"female"` or `"male"` for estimated general
combining abilities.

For a purely additive model the distinction between `"gv"` and `"bv"` is
cosmetic: both use the same marker effects and differ only in the
intercept, so they are perfectly correlated and select the same
individuals. The distinction matters once dominance is in the model,
because then the genetic value and the breeding value are different
quantities and rank individuals differently.

`append` controls whether the new estimates replace the contents of the
`ebv` slot or are added as an extra column. Appending is how you carry
several estimates at once — a genetic value and a breeding value from
the same fit, or estimates for several traits to be combined in an
index.

### The reference population

An estimated breeding value depends on allele frequencies, because an
average effect does. By default `setEBV` uses the frequencies implied by
the fitted model, which are those of the training population. The
`targetPop` argument overrides this, recomputing the average effects
$\alpha = a + d(q - p)$ at the allele frequencies of the population you
supply:

```{r}
candidates = setEBV(candidates, ans, value = "bv", targetPop = candidates,
                    append = TRUE, simParam = SP)

cor(ebv(candidates)[, 1], ebv(candidates)[, 2])
```

For an additive model this changes nothing but the intercept, as the
correlation of one shows. For a model with dominance it genuinely
changes the ranking, and it is the right thing to do when the population
being selected has drifted away from the population the model was
trained on. In a hybrid program it is how a parent is evaluated against
the *opposite* pool, by passing that pool as `targetPop`.

### Measuring accuracy

Accuracy is the correlation between the estimate and the thing it is
meant to estimate. Which "thing" depends on the decision. If you are
selecting parents, breeding value is the relevant target; if you are
selecting a variety to release, genetic value is.

```{r}
cor(ebv(candidates)[, 1], gv(candidates)[, 1])
cor(ebv(candidates)[, 1], bv(candidates, simParam = SP)[, 1])
```

Two mistakes are easy to make. The first is measuring accuracy in the
training population. Those individuals contributed to the fit, so the
correlation there is inflated and says little about performance on the
individuals you actually have to select:

```{r}
trainPop = setEBV(trainPop, ans, simParam = SP)

# Inflated: these individuals trained the model
cor(ebv(trainPop)[, 1], gv(trainPop)[, 1])
```

The second is comparing a correlation against genetic value with one
against breeding value and treating the difference as a property of the
model. With dominance in the trait the two targets differ, and the gap
between them is a property of the trait, not of the prediction.

## Practical matters

### Choosing the markers

By default the models use the first SNP chip. A different chip is
selected with `snpChip`, and `useQtl=TRUE` substitutes a trait's QTL for
the chip. The QTL option is a diagnostic: it gives an upper bound on
what a model can achieve, because it removes the imperfect linkage
between markers and causal loci. Be aware that when `useQtl=TRUE` the
`snpChip` argument switches meaning and selects *which trait's* QTL to
use, so those QTL need not be the ones underlying the response you are
modeling.

### Size and memory

The cost of these models is driven by the number of training records and
the number of markers, and the balance between the two decides which
implementation to use. `RRBLUPMemUse` estimates the memory required, and
takes the model as the name of the function that fits it:

```{r}
RRBLUPMemUse(nInd = 5000, nMarker = 2000, model = "fastRRBLUP")
RRBLUPMemUse(nInd = 5000, nMarker = 2000, model = "RRBLUP")
RRBLUPMemUse(nInd = 5000, nMarker = 2000, model = "RRBLUP_SCA")
```

The estimate covers the matrices the solvers hold at their peak, not the
population object or the rest of the session, so it is a lower bound on
what the simulation needs. It is most useful for spotting the point at
which a training population outgrows the machine.

## What the models fit

This section is the background to the choices above, and can be skipped.

All of these models are ridge regressions on marker genotypes,
$$y = X\beta + Mu + e$$
where $M$ holds the marker covariates, the marker effects $u$ are
treated as random with a common variance $V_u$, and $\beta$ holds the
fixed effects discussed in the next section. The shrinkage applied to
the marker effects is set by the ratio $V_e / V_u$, so the variance
components have to be estimated before the effects can be.

Markers are coded on the same scale AlphaSimR uses for trait effects:
for a diploid, the additive covariate takes the values −1, 0 and 1 for
the three genotypes, and the dominance covariate takes the value 1 for
the heterozygote and 0 otherwise. Estimated effects are therefore
directly comparable with the true QTL effects in `SP$traits`, which is
convenient for checking a model against the truth.

A fitted model is returned as an `RRsol-class` object with six slots:

```{r}
class(ans)

ans@Vu
ans@Ve
```

`Vu` and `Ve` hold the estimated marker and residual variances. The
remaining four — `gv`, `bv`, `female` and `male` — hold *trait objects*,
the same kind of object `SimParam` uses to define a real trait. This is
what makes a solution portable: a fitted model is stored as an estimated
trait, so applying it to a new population is the same operation as
evaluating a genetic value, and it works on any population carrying the
same markers.

Which slots are filled depends on the model. `RRBLUP` and `RRBLUP_D`
fill `gv` and `bv`; the GCA and SCA models fill `gv`, `female` and
`male` and leave `bv` empty. Asking `setEBV` for a value the model did
not produce raises an error rather than returning something misleading.

```{r}
length(ans@gv)
length(ans@bv)
length(ans@female)
```

### How the equations are solved

The three implementations differ only in arithmetic, not in the model.

The **unnumbered** functions use the approach of Kang and colleagues,
sometimes called EMMA: the problem is transformed by an
eigendecomposition of a matrix whose dimension is the number of
*records*, and the variance components are then found by maximizing the
restricted likelihood along a single dimension. This is fast and
reliable, but it forms an $n \times n$ matrix.

The **numbered** functions set up Henderson's mixed model equations
directly. The coefficient matrix there has dimension equal to the number
of fixed effects plus the number of *markers*, and the variance
components are found with an EM algorithm. So if the number of markers
is appreciably smaller than the number of records, the numbered
functions use less memory; if the number of markers is larger, which is
the common case, the unnumbered functions use less.

**`fastRRBLUP`** forms no square matrix at all. It solves the mixed
model equations by preconditioned conjugate gradient, iterating over the
genotypes rather than over a stored coefficient matrix
[@stranden_1999]. Neither the coefficient matrix nor a numeric copy of
the genotypes is ever formed, so its memory cost stays close to the cost
of the genotypes themselves, and both halves of each iteration are
shared across `nThreads`.

It also differs in how it gets its variance components. REML needs a
matrix decomposition costing the square of the number of records in
memory and their cube in time, which is exactly what `fastRRBLUP` exists
to avoid, so it estimates the variance components from a random subset
of the records. `nSubsample` sets the size of that subset and defaults
to 5000. Variance components are nuisance parameters here, estimated far
more precisely than a breeding program needs, so a few thousand records
give values good enough to shrink with — and the marker effects
themselves are always solved for using every record. Supplying `Vu` and
`Ve` skips the estimation entirely.

## Fixed effects

The models fit an intercept by default, and that is all most simulations
need. The `fixEff` slot of a population labels each record with a group,
and the models fit those groups as fixed effects. The label is set by
the `fixEff` argument of `setPheno` and is carried by the population
from then on:

```{r}
batch1 = setPheno(pop[1:200], fixEff = 1L, simParam = SP)
batch2 = setPheno(pop[201:400], fixEff = 2L, simParam = SP)

table(c(batch1, batch2)@fixEff)
```

Internally each model applies `as.integer(factor(pop@fixEff))`, so the
labels need not be consecutive or start at one; only the distinct values
matter. The design matrix uses sum-to-zero coding with an intercept in
the first column, so a population with a single `fixEff` level reduces
to fitting an intercept alone.

It is worth being explicit about what `fixEff` does *not* do. It does
not add anything to the phenotype. AlphaSimR will not shift the mean of
`batch2` simply because it was labeled `2`. The label lets a model
absorb differences that *you* created. If you generate all records the
same way, the fitted fixed effects estimate nothing.

This detail matters in one situation in particular: **combining
populations phenotyped in different environments into a single training
population, when the trait has a genotype-by-environment component.**
A trait created with `varGxE` and `varEnv` — through `addTraitAG`,
`addTraitADG` and their relatives — has an environment effect that
shifts the mean of every record taken in that environment, with the
environment drawn by the `p` argument of `setPheno`. Records from a
first-year trial and a second-year trial are then on genuinely different
scales. Pooling them without a label leaves that difference in the
residual, where it inflates $V_e$, shrinks the marker effects too hard
and lowers accuracy. Labelling each trial with its own `fixEff` lets the
model absorb the shift:

```{r eval=FALSE}
# Each trial phenotyped in its own environment, and labelled as such
PYT = setPheno(PYT, reps = repPYT, p = pYear, fixEff = year, simParam = SP)
AYT = setPheno(AYT, reps = repAYT, p = pYear, fixEff = year, simParam = SP)

trainPop = c(trainPop, PYT, AYT)
```

The sliding-window training population described earlier is exactly this
case, since it spans several years of trials. For a trait with no GxE
component, and for records generated identically, the labels can be left
alone.

## Using external software

The AlphaSimR models cover ridge regression and its close relatives. For
anything else — a Bayesian model, a GBLUP with a particular relationship
matrix, a machine learning method — export the genotypes and do the
prediction elsewhere.

Genotypes come out as an ordinary matrix, with individuals in rows and
markers in columns, coded as allele dosages:

```{r}
geno = pullSnpGeno(trainPop, snpChip = 1, simParam = SP)

dim(geno)
geno[1:5, 1:5]
```

The matching marker map, which most external software will want, comes
from `getSnpMap`:

```{r}
head(getSnpMap(snpChip = 1, simParam = SP))
```

`pullQtlGeno` and `pullSegSiteGeno` do the same for QTL and for all
segregating sites, and the `pull*Haplo` family returns haplotypes rather
than genotypes. For software that reads files rather than matrices,
`writePlink` writes a population in PLINK format and `writeRecords`
writes phenotype records.

Bringing predictions back is a matter of writing them into the `ebv`
slot directly. `setEBV` is a convenience for AlphaSimR's own solutions,
not a requirement:

```{r eval=FALSE}
# 'pred' is a vector of predictions from an external package,
# in the same order as the individuals in candidates
candidates@ebv = as.matrix(pred)
```

The slot must be a numeric matrix with one row per individual; any
number of columns is allowed, and the selection functions index it with
their `trait` argument. Once it is filled, the population behaves
exactly as it would after `setEBV`, and every selection function with
`use="ebv"` will work.

Two R packages worth mentioning here are `rrBLUP` and `sommer`
[@sommer], both of which fit mixed models with arbitrary covariance
structures and accept the matrices produced above without modification.
`EMMREML` [@EMMREML] is another option for models with known covariance
structures.

## References
