---
title: "Genomic Selection in AlphaSimR"
author: "Chris Gaynor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: AlphaSimR.bib
vignette: >
  %\VignetteIndexEntry{Genomic Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(AlphaSimR)
```

## Introduction

Genomic selection is selection on a prediction of an individual's genetic
merit made from its marker genotypes. AlphaSimR provides a family of
ridge regression BLUP (RR-BLUP) models for making these predictions
[@AlphaSim; @gaynor_2017], following the marker-effect approach
introduced by Meuwissen and colleagues. The models are fitted in C++ and
are intended to be re-fitted many times over the course of a simulation,
so they are deliberately narrow in scope: they are fast, they take their
data directly from a population object, and they return an object that
AlphaSimR knows how to turn back into estimated values for any population
you like.

Two earlier articles are assumed here. The article on the genome explains
how QTL and SNP chips are defined and how genotypes are stored and
pulled. The article on quantitative genetics explains the distinction
between a genetic value and a breeding value, and explains why a breeding
value is a property of a *reference population* rather than of an
individual on its own. That distinction reappears below, because it
determines what `setEBV` gives you.

Every genomic selection analysis in AlphaSimR has the same three steps:

1. Assemble a training population that carries both phenotypes and
   genotypes.
2. Fit a model to it, producing an object of `RRsol-class`.
3. Apply that solution to a population with `setEBV`, which fills the
   population's `ebv` slot.

Selection then proceeds with the ordinary selection functions, using
`use="ebv"`.

## A worked example

The example below is used throughout the article.

```{r}
set.seed(1234)

founderPop = runMacs(nInd = 500, nChr = 10, segSites = 100)

SP = SimParam$new(founderPop)
```
```{r include=FALSE}
SP$nThreads = 1L
```
```{r}
# Keep the QTL and the SNP chip on separate segregating sites
SP$restrSegSites(minQtlPerChr = 20, minSnpPerChr = 50, overlap = FALSE)

SP$addTraitAD(nQtlPerChr = 20, meanDD = 0.4)
SP$addSnpChip(nSnpPerChr = 50)
SP$setVarE(h2 = 0.4)

pop = newPop(founderPop, simParam = SP)
```

Note the use of `runMacs` rather than `quickHaplo`. Genomic prediction
works because markers are in linkage disequilibrium with QTL, and
`quickHaplo` samples haplotypes at linkage equilibrium. A simulation
built on `quickHaplo` will show almost no predictive ability unless the
QTL themselves are used as markers, which makes it a poor choice for
demonstrating or benchmarking genomic selection.

`restrSegSites` is used here to force the QTL and the SNP chip onto
different segregating sites. Without it the two sets are sampled
independently and may overlap, which quietly inflates accuracy.

## Building a training population

A training population is an ordinary `Pop-class` object. There is no
special class and no separate registration step; a model simply reads the
response and the genotypes out of whatever population you hand it. The
requirements are only that the population carries a value to train on and
that the SNP chip you intend to use was defined before the population was
created.

The response is chosen with the `use` argument, which accepts `"pheno"`
(the default), `"gv"`, `"bv"`, `"ebv"`, or `"rand"`. Training on
phenotypes is the realistic choice. Training on `"gv"` or `"bv"` is
useful for diagnosis, because it removes environmental noise and shows
what the model could achieve if phenotyping were perfect.

Training populations are usually assembled from several sources, and the
natural tool for that is `c()`:

```{r}
# Two batches of records, phenotyped separately
batch1 = setPheno(pop[1:200], reps = 1, fixEff = 1L, simParam = SP)
batch2 = setPheno(pop[201:400], reps = 3, fixEff = 2L, simParam = SP)

trainPop = c(batch1, batch2)

# Held back for validation
predPop = pop[401:500]

nInd(trainPop)
```

### Fixed effects

Records assembled from different trials, years or stages of a breeding
program are rarely on a common scale. The `fixEff` slot of a population
labels each record with the group it came from, and the models fit those
groups as fixed effects.

```{r}
table(trainPop@fixEff)
```

The label is set by the `fixEff` argument of `setPheno` and is carried by
the population from then on. Internally each model applies
`as.integer(factor(pop@fixEff))`, so the labels need not be consecutive
or start at one; only the distinct values matter. The design matrix uses
sum-to-zero coding with an intercept in the first column, so a population
with a single `fixEff` level reduces to fitting an intercept alone, which
is what happens by default.

It is worth being explicit about what `fixEff` does *not* do. It does not
add anything to the phenotype. AlphaSimR will not shift the mean of
`batch2` simply because it was labelled `2`. The label exists so that a
model can absorb differences that *you* created, for example by
phenotyping the two batches with different error variances, different
numbers of replicates, or different environmental covariates. If you
generate all records the same way, the fitted fixed effects will simply
be estimates of nothing.

### Choosing the markers

By default the models use the first SNP chip. A different chip is
selected with `snpChip`, and `useQtl=TRUE` substitutes a trait's QTL for
the chip. The QTL option is a diagnostic: it gives an upper bound on what
a model can achieve, because it removes the imperfect linkage between
markers and causal loci. Be aware that when `useQtl=TRUE` the `snpChip`
argument switches meaning and selects *which trait's* QTL to use, so
those QTL need not be the ones underlying the response you are modelling.

### Size and memory

The cost of these models is driven by the number of training records and
the number of markers, and the balance between the two decides which
implementation to use. `RRBLUPMemUse` gives a rough estimate of the
memory required:

```{r}
RRBLUPMemUse(nInd = 2000, nMarker = 10000)
RRBLUPMemUse(nInd = 2000, nMarker = 10000, model = "SCA")
```

The estimate is approximate and, as the help page warns, can be an
underestimate. It is most useful for spotting the point at which a
training population becomes too large for the machine, which is normally
driven by the terms in $n^2$.

## Fitting a model

The simplest model is `RRBLUP`, which fits additive marker effects:

```{r}
ans = RRBLUP(trainPop, simParam = SP)

class(ans)
```

The result is an object of `RRsol-class` with six slots. `Vu` and `Ve`
hold the estimated marker and residual variances:

```{r}
ans@Vu
ans@Ve
```

The remaining four slots — `gv`, `bv`, `female` and `male` — hold *trait
objects*, the same kind of object that `SimParam` uses to define a real
trait. This is the mechanism that makes the solution portable. A fitted
model is stored as an estimated trait, so applying it to a new population
is exactly the same operation as evaluating a genetic value, and it can
be applied to any population that carries the same markers.

Which of the four slots are filled depends on the model. `RRBLUP` fills
`gv` and `bv`; the GCA and SCA models fill `gv`, `female` and `male` but
leave `bv` empty. Asking `setEBV` for a value the model did not produce
raises an error rather than returning something misleading.

```{r}
length(ans@gv)
length(ans@bv)
length(ans@female)
```

Markers are coded on the same scale that AlphaSimR uses for trait
effects: for a diploid, the additive covariate takes the values −1, 0 and
1 for the three genotypes, and the dominance covariate takes the value 1
for the heterozygote and 0 otherwise. Estimated effects are therefore
directly comparable with the true QTL effects returned by `SP$traits`,
which is convenient when you want to check a model against the truth.

## Getting GEBVs

`setEBV` applies a solution to a population and writes the result into
the population's `ebv` slot:

```{r}
predPop = setEBV(predPop, ans, simParam = SP)

head(ebv(predPop))
```

The `value` argument chooses what is estimated. The default is `"gv"`, an
estimated genetic value. The alternatives are `"bv"` for an estimated
breeding value and `"female"` or `"male"` for estimated general combining
abilities in the two pools of a hybrid program.

For a purely additive model the distinction between `"gv"` and `"bv"` is
cosmetic: both use the same marker effects and differ only in the
intercept, so they are perfectly correlated and select the same
individuals. The distinction matters once dominance is in the model,
because then the genetic value and the breeding value are different
quantities.

`append` controls whether the new estimates replace the contents of the
`ebv` slot or are added as an extra column. Appending is how you carry
several estimates at once, for example a genetic value and a breeding
value from the same fit, or estimates for several traits to be combined
in an index.

### The reference population

An estimated breeding value depends on allele frequencies, because an
average effect does. By default `setEBV` uses the frequencies implied by
the fitted model, which are those of the training population. The
`targetPop` argument overrides this, recomputing the average effects
$\alpha = a + d(q - p)$ at the allele frequencies of the population you
supply:

```{r}
# Breeding values relative to the prediction population
predPop = setEBV(predPop, ans, value = "bv", targetPop = predPop,
                 append = TRUE, simParam = SP)

cor(ebv(predPop)[, 1], ebv(predPop)[, 2])
```

For an additive model this changes nothing but the intercept, as the
correlation of one shows. For a model with dominance it genuinely changes
the ranking, and it is the right thing to do when the population you are
selecting in has drifted away from the population the model was trained
on. The quantitative genetics article discusses why.

### Measuring accuracy

Accuracy is the correlation between the estimate and the thing it is
meant to estimate. Which "thing" depends on the decision being made. If
you are selecting parents, breeding value is the relevant target; if you
are selecting a variety to be released, genetic value is.

```{r}
# Accuracy in the prediction population
cor(ebv(predPop)[, 1], gv(predPop)[, 1])
cor(ebv(predPop)[, 1], bv(predPop, simParam = SP)[, 1])
```

Two things are easy to get wrong here. The first is measuring accuracy in
the training population. Those individuals contributed to the fit, so the
correlation there is inflated and says little about how the model will
perform on the individuals you actually have to select:

```{r}
trainPop = setEBV(trainPop, ans, simParam = SP)

# Inflated: these individuals trained the model
cor(ebv(trainPop)[, 1], gv(trainPop)[, 1])
```

The second is comparing a correlation against genetic value with one
against breeding value and treating the difference as a property of the
model. With dominance in the trait the two targets differ, and the gap
between them is a property of the trait, not of the prediction.

## Making selections

Once the `ebv` slot is filled, genomic selection is ordinary selection
with `use="ebv"`. Every selection function takes the argument:

```{r}
selected = selectInd(trainPop, nInd = 50, use = "ebv", simParam = SP)

meanG(selected)
meanG(trainPop)
```

`selectCross` combines selection and crossing in one call, which is the
usual form inside a breeding cycle:

```{r}
newPop = selectCross(trainPop, nInd = 50, nCrosses = 200,
                     use = "ebv", simParam = SP)
```

A full genomic selection cycle therefore has four lines: phenotype,
fit, predict, select. The only real design question is how often the
model is re-trained and what goes into the training population when it
is. Re-fitting every cycle on an accumulating training population is
the common choice, but it is also the expensive one, and the next
section is largely about what to do when it becomes too expensive.

## The models

AlphaSimR offers five model structures, each in two implementations.

| Model | Effects fitted | Slots filled |
|---|---|---|
| `RRBLUP` | additive | `gv`, `bv` |
| `RRBLUP_D` | additive and dominance | `gv`, `bv` |
| `RRBLUP_GCA` | separate additive effects per pool | `gv`, `female`, `male` |
| `RRBLUP_SCA` | separate additive effects per pool, plus dominance | `gv`, `female`, `male` |
| `fastRRBLUP` | additive | `gv`, `bv` |

`RRBLUP` is the default choice and the only one that fits several traits
at once. Supplying more than one trait fits a multivariate model, which
borrows information across correlated traits at the cost of estimating a
full covariance matrix.

`RRBLUP_D` adds a dominance covariate for each marker. It is the model to
use when the trait was defined with dominance and you care about the
genetic value rather than only the breeding value — predicting the
performance of a cross, for instance, rather than the merit of a parent
[@xiang_2016].

`RRBLUP_GCA` and `RRBLUP_SCA` are for hybrid programs with two heterotic
pools. Rather than one set of additive effects, they estimate a set for
each pool, which allows an allele to be worth something different
depending on which parent contributed it. `RRBLUP_SCA` adds a dominance
term on top, so that specific combining ability can be predicted as well
as general combining ability [@Massman13]. Both fill the `female` and
`male` slots with pool-specific solutions and leave `bv` empty; ask for
`value="female"` or `value="male"` to get GCA estimates, and `value="gv"`
to predict the performance of a specific cross. In practice these models
are paired with `setPhenoGCA`, which phenotypes candidate parents through
testcrosses, and with `hybridCross` for producing the crosses themselves.

@dejong_2023 compared these models for the specific problem of selecting
inbred parents on predicted GCA in a simulated maize program, and found
that models fitting pool-specific effects together with dominance gave
the best fit and the highest genetic gain. They also found that the
ranking of models depends on marker density, with the models differing
more in genetic gain at low density and more in heterosis at high
density, which is a useful warning against assuming one model is
uniformly best.

`fastRRBLUP` is the exception in the list. It does not estimate variance
components at all — `Vu` and `Ve` must be supplied, and reasonable
starting values are chosen from the simulation parameters if they are
not — and it solves the resulting system by Gauss–Seidel iteration over
the markers. It never forms a large matrix, so its memory cost is close
to the cost of the genotypes themselves. It is the model to reach for
when the training population is large enough that the others will not
fit in memory.

### Model 1 versus model 2

Every model above has a numbered counterpart: `RRBLUP2`, `RRBLUP_D2`,
`RRBLUP_GCA2`, `RRBLUP_SCA2`. These fit exactly the same models. They
differ only in how the equations are solved, and the difference is one of
memory and speed rather than of results.

The unnumbered functions use the approach of Kang and colleagues,
sometimes called EMMA: the problem is transformed by an eigendecomposition
of a matrix whose dimension is the number of *records*, and the variance
components are then found by maximising the restricted likelihood along a
single dimension. This is fast and reliable, but it forms an $n \times n$
matrix.

The numbered functions instead set up Henderson's mixed model equations
directly. The coefficient matrix there has dimension equal to the number
of fixed effects plus the number of *markers*, and the variance
components are found with an EM algorithm. So:

- If the number of markers is appreciably smaller than the number of
  records, the numbered functions use less memory.
- If the number of markers is larger, which is the common case, the
  unnumbered functions use less memory.
- The EM algorithm is slower and less reliable than the EMMA approach for
  estimating variance components. `maxIter` defaults to a value that is
  usually too small to converge, and a warning is issued when it does not.

The numbered functions come into their own when the variance components
do not need to be estimated. Setting `useEM=FALSE` and supplying `Vu` and
`Ve` reduces the work to a single solve of the mixed model equations,
which is quick. A common pattern is to fit `RRBLUP` once at the start of
a simulation, keep its `Vu` and `Ve`, and use `RRBLUP2` with those values
for every subsequent re-training:

```{r}
ans2 = RRBLUP2(trainPop, Vu = ans@Vu[1, 1], Ve = ans@Ve[1, 1],
               useEM = FALSE, simParam = SP)

predPop = setEBV(predPop, ans2, simParam = SP)

cor(ebv(predPop)[, 1], gv(predPop)[, 1])
```

Whether holding the variance components fixed is acceptable depends on
how far the population drifts over the course of the simulation, and that
is a question only the simulation itself can answer. It is worth checking
against a full re-fit before relying on it.

## Using external software

The AlphaSimR models cover ridge regression and its close relatives. For
anything else — a Bayesian model, a GBLUP with a particular relationship
matrix, a machine learning method — the genotypes are exported and the
prediction is done elsewhere.

Genotypes come out as an ordinary matrix, with individuals in rows and
markers in columns, coded as allele dosages:

```{r}
geno = pullSnpGeno(trainPop, snpChip = 1, simParam = SP)

dim(geno)
geno[1:5, 1:5]
```

The matching marker map, which most external software will want, comes
from `getSnpMap`:

```{r}
head(getSnpMap(snpChip = 1, simParam = SP))
```

`pullQtlGeno` and `pullSegSiteGeno` do the same for QTL and for all
segregating sites, and the `pull*Haplo` family returns haplotypes rather
than genotypes. For software that reads files rather than matrices,
`writePlink` writes a population in PLINK format and `writeRecords`
writes phenotype records.

Bringing the predictions back is a matter of writing them into the `ebv`
slot directly. `setEBV` is a convenience for AlphaSimR's own solutions,
not a requirement:

```{r eval=FALSE}
# 'pred' is a vector of predictions from an external package,
# in the same order as the individuals in predPop
predPop@ebv = as.matrix(pred)
```

The slot must be a numeric matrix with one row per individual; any number
of columns is allowed, and the selection functions index it with their
`trait` argument. Once it is filled, the population behaves exactly as it
would after `setEBV`, and every selection function with `use="ebv"` will
work.

Two R packages worth mentioning in this context are `rrBLUP` and
`sommer` [@sommer], both of which fit mixed models with arbitrary
covariance structures and both of which accept the matrices produced
above without modification. `EMMREML` [@EMMREML] is another option for
models with known covariance structures.

## References
