---
title: "Quantitative Genetics in AlphaSimR"
author: "Gregor Gorjanc and Chris Gaynor"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: '`r normalizePath(if (file.exists("../inst/REFERENCES.bib")) "../inst/REFERENCES.bib" else system.file("REFERENCES.bib", package = "AlphaSimR", mustWork = TRUE), winslash = "/", mustWork = TRUE)`'
vignette: >
  %\VignetteIndexEntry{Quantitative Genetics in AlphaSimR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(AlphaSimR)
```

## Introduction and reading guide

This vignette explains how AlphaSimR constructs genetic values from genotypic effects,
decomposes them into population-specific components, and calculates their variances.
It is expected that the reader is familiar with the background theory
[@falconer1996Introduction; @lynch1998Genetics].
Also, the companion `vignette("traits")` introduces trait specification and
should be read first.
For an application to cross performance involving multiple populations (pools / breeds),
`vignette("CombiningAbility")` demonstrates general and specific combining ability (GCA and SCA).

<!-- TODO: Add the companion imprinting vignette here when it is available. -->

| If you want to understand... | Start with... |
|:--|:--|
| Difference between two common parameterizations of genetic values | [Two parameterizations of genetic values](#two-parameterizations) |
| How genotypes and trait effects determine genetic values | [Genotypic effects and genetic values](#genotypic-values) |
| Why breeding values and non-additive deviations depend on a population | [Breeding values and non-additive deviations](#breeding-values) |
| How genetic variance differs from genic variance | [Genetic and genic variances](#genetic-variances) |
| How inbreeding affects genetic variance | [Departures from Hardy-Weinberg proportions](#hardy-weinberg-departures) |
| How selection affects genetic variance | [Linkage disequilibrium and the Bulmer effect](#disequilibrium) |
| What changes for polyploids | [Polyploids and digenic dominance](#polyploids) |
| How AlphaSimR implementation relates to classical quantitative genetics | [Relationship to classical theory](#classical-theory) |

The examples build on earlier code:
use the links to navigate when reading,
but run the examples in order.

All of the quantitative genetics quantities described below are computed by `genParam`.
User-friendly functions, such as `bv`, internally call `genParam`.

For a more detailed numerical walkthrough,
see the `genParam` unit tests in `tests/testthat/test-genParam.R`
in the AlphaSimR source package or git repository.

## Two parameterizations of genetic values {#two-parameterizations}

There are two commonly used parameterizations of genetic values,
the genotypic parameterization and breeding value parameterization, and
they are a constant source of confusion.
They describe the same genetic values and are hence equivalent from that point of view,
but divide the genetic values into different components that differ in their meaning.
The key distinction is how those components depend on the population.

The *genotypic parameterization* constructs genetic values from genotypes,
genotypic effects, and a trait intercept.
Once the effects and intercept are specified,
the genetic value assigned to a genotype does not depend
on allele or genotype frequencies.
This is sometimes called the "functional" or "biological" model,
but note that this model is a very simplified representation of trait biology
to enable mathematical analysis.

The *breeding value parameterization* decomposes genetic values into
a population mean, additive genetic (breeding) values, and
non-additive deviations.
The non-additive effects considered in this vignette are dominance and epistasis.
These components depend on allele and genotype frequencies in a population.
The same individual can therefore have
different breeding values and non-additive deviations in different populations,
even when their genetic value is unchanged.
This is sometimes called the "statistical" model,
because we obtain breeding values and deviations from fitting a statistical model
to a specific population.

Both parameterizations can be used for simulation, theoretical calculations, and
estimation from data.
The breeding value parameterization underlies much of quantitative genetics,
evolution, and breeding theory [@falconer1996Introduction; @lynch1998Genetics] and
is the main focus here.
In the following, we first describe the genotypic parameterization,
then derive breeding values and non-additive deviations and their
corresponding variances.

Note that ADEG symbols can denote different quantities in the two parameterizations,
so symbol definitions should be read carefully.

## Genotypic effects and genetic values {#genotypic-values}

### Genotype coding and the genetic value model

AlphaSimR builds the genetic value of an individual from biallelic loci
using scaled genotype dosages following the *genotypic parameterization*.
For a locus with raw dosage $x$ (the number of copies of the "1" allele)
in an organism of ploidy $P$,
the scaled additive and dominance dosages are respectively:

$$x_A = \left( x - \frac{P}{2} \right)\left( \frac{2}{P} \right)
\qquad\text{and}\qquad
x_D = x \left( P - x \right) \left( \frac{2}{P} \right)^{2},$$

and the genetic value of an individual is:

$$G(x) = \mu_0 + \sum_i x_{A_i} a_i + \sum_i x_{D_i} d_i +
\sum_{(i,j)} x_{A_i} x_{A_j} e_{(i,j)},$$

where $\mu_0$ is the trait intercept,
calibrated to achieve the requested founder mean and not generally equal to that mean,
$a_i$ and $d_i$ are additive and dominance effects at locus $i$, and
$e_{(i,j)}$ is an additive-by-additive epistatic effect between the pair of loci $i$ and $j$.
The additive and dominance sums are over the causal (quantitative trait) loci (QTL),
while the epistatic sum is over the *pairs* of QTL.

AlphaSimR assigns each QTL to exactly one epistatic pair,
so these pairs are disjoint and the number of pairs is half the number of QTL.
This means the two-locus calculations described below
can be carried out one pair at a time and summed.
Note that the only modeled epistasis is additive-by-additive.
Additive-by-dominance or dominance-by-dominance epistasis are not supported.

Note that genetic values in AlphaSimR include the trait intercept $\mu_0$,
while standard literature often removes it (or the trait mean of a population)
[@fisher1919Correlation, @falconer1996Introduction, @lynch1998Genetics].
While this difference only shifts the "baseline" of genetic values,
it is convenient for seamless comparisons across populations and time.
Since there are several "baseline"-related quantities in AlphaSimR,
we next explain these concretely.

### Founder mean, trait intercept, and population mean {#means-and-intercept}

It is useful to distinguish between
the mean requested when specifying a trait from
the intercept used to calculate genetic values, and
a population mean.
The intercept sets a fixed baseline,
while the population mean also depends on the genotypes in the population.
Three quantities therefore need to be distinguished:

| Quantity | Symbol | R quantity |
|:--|:--|:--|
| Requested founder mean | $m_{\mathrm{target}}$ | `mean` argument to `addTrait*()` |
| Trait intercept | $\mu_0$ | `SP$traits[[t]]@intercept` and `genParam(pop)$gv_mu[t]` for trait `t` |
| Current population mean genetic value | $\bar{G}$ | `meanG(pop)` and `genParam(pop)$mu` |

These are related in the following way.
Let $C_i$ be the sum of individual $i$'s contributions from genotypic effects.
Then:

$$G_i = \mu_0 + C_i \qquad \text{and} \qquad \bar{G} = \mu_0 + \bar{C}.$$

After scaling the effects to achieve the requested founder variance,
AlphaSimR chooses the trait intercept as:

$$\mu_0 = m_{\mathrm{target}} - \bar{C}_{\mathrm{founders}}.$$

Thus the founder mean equals the requested mean,
but the intercept generally differs from it.
With the trait definition held fixed,
the intercept remains constant
while the population mean can change as genotype frequencies change.
The corresponding identity for the quantities returned by `genParam()`
(as shown below) is `mu = gv_mu + colMeans(gv_a + gv_d + gv_aa)`, where
`mu` is current population mean genetic value $\bar{G}$,
`gv_mu` is trait intercept $\mu_0$, and
`colMeans(gv_a + gv_d + gv_aa)` is the mean of contributions from genotypic effects $\bar{C}$.

### Worked example: one locus

We now illustrate how the intercept and genotypic effects combine
to give genetic values at one locus.
First we simulate founder genomes and
specify a trait with one QTL, a requested founder mean of 10, and
additive (default) and dominance genotypic effects.

```{r}
# Founders and a trait
founderPop = quickHaplo(nInd = 100, nChr = 1, segSites = 1)
SP = SimParam$new(founderPop)
SP$addTraitAD(nQtlPerChr = 1, mean = 10, meanDD = 0.75)

# Parameters of the genotypic parameterization
SP$traits[[1]]@intercept
SP$traits[[1]]@addEff
SP$traits[[1]]@domEff
```

Now we use this setup in a population and
show the *genotypic parameterization of genetic values*.

```{r}
# Simulate a population
pop = newPop(founderPop)

# QTL genotypes
popQtlGenos = pullQtlGeno(pop)
x_A = popQtlGenos - 1
x_D = popQtlGenos * (2 - popQtlGenos)

# Genotypic parameterization of genetic values
tmp = cbind(popQtlGenos, gv(pop),
            SP$traits[[1]]@intercept,
            x_A, x_A * SP$traits[[1]]@addEff,
            x_D, x_D * SP$traits[[1]]@domEff)
colnames(tmp) = c("genotype", "gv", "mu0", "x_A", "x_A_a", "x_D", "x_D_d")
head(tmp)
# ... genetic value for the individual i
id = 1 # change to inspect other individuals
tmp[id, "gv"]
# ... and its genotypic parameterization
tmp[id, "mu0"] + tmp[id, "x_A_a"] + tmp[id, "x_D_d"]
```

## Breeding values and non-additive deviations {#breeding-values}

### Average effects and breeding values

Under the *breeding value parameterization*
AlphaSimR decomposes the genetic value of an individual with genotype $x$ into
the population mean $\bar{G}$, breeding value $A(x)$, and non-additive deviations $N(x)$:

$$G(x) = \bar{G} + A(x) + N(x).$$

Here $\bar{G}$ is the mean genetic value in the current population,
rather than the fixed trait intercept $\mu_0$ introduced
[above](#means-and-intercept).

The breeding value of an individual is the part of its genetic value that
it can transmit to offspring under random mating in the population
[@falconer1996Introduction, @falconer1985Note].
Formally,
it is the least-squares projection of
the genetic-value deviation from the population mean onto
centred allele dosage in the population,
and the coefficient of that projection is
the average effect of allele substitution in the population,
usually denoted as $\alpha$.

AlphaSimR calculates these regression coefficients using the observed genotype frequencies,
without requiring Hardy–Weinberg proportions.
When genotype frequencies depart from these proportions,
the resulting additive genetic values remain valid
as a statistical description of the current population,
but need not equal twice the expected deviation of
an individual’s offspring mean from the population mean under random mating [@falconer1985Note].

We now describe the $\alpha$ and breeding values in more detail and
later turn attention to non-additive deviations.

Consider a single locus in a population where genotype dosage $x$ occurs
with frequency $f(x)$.
Let $\bar{x} = \sum_x f(x)\, x$ be the mean dosage and
$G(x)$ the genetic value of dosage $x$.
The average effect of allele substitution is
the regression coefficient of genetic value on dosage:

$$\alpha = \frac{\sum_x f(x)\,\big(G(x) - \bar{G}\big)\big(x -
\bar{x}\big)}{\sum_x f(x)\,\big(x - \bar{x}\big)^{2}},$$

and the breeding value of an individual with dosage $x$ is:

$$A(x) = \big(x - \bar{x}\big)\, \alpha .$$

A breeding value of an individual for a trait is the sum of $A(x)$ over all QTL.

Three features of this definition deserve emphasis.
First, $\alpha$ and hence also $A(x)$ are not a property of the locus alone.
They depend on allele and genotype frequencies in a population
and are hence population specific.
Second, when there are no non-additive genotypic effects,
$G(x)$ is linear in $x$ and the regression is exact.
With the scaled dosage defined above,
the slope on raw dosage is $\alpha = 2a/P$,
which equals the additive genotypic effect $a$ for diploids.
Breeding values then account for all of the genetic variation.
Third, breeding values are centered on the population mean by construction,
since $\sum_x f(x)(x - \bar{x}) = 0$.

AlphaSimR computes two versions of $\alpha$ for every locus using `genParam` function.
The first uses the genotype frequencies actually observed in the population and
is returned as `alpha`.
The second uses the genotype frequencies expected under Hardy-Weinberg equilibrium
at the observed allele frequency, and is returned as `alpha_HW`.
Writing $p = \bar{x}/P$ for the allele frequency,
the Hardy-Weinberg genotype frequencies for ploidy $P$ are:

$$f_{HW}(x) = \binom{P}{x} p^{x} (1-p)^{P-x},$$

which is the familiar binomial expansion.
For $P = 2$ we obtain the usual genotype frequencies of $p^2$, $2pq$, and $q^2$.
The two versions of $\alpha$ can differ
when the population departs from Hardy-Weinberg proportions,
for example through inbreeding or non-random mating.
The difference between these settings is what AlphaSimR reports as
the non-random-mating components of variance.

We now return to demonstration.
We first demonstrate genotype and allele frequencies,
followed by $\alpha$ and *breeding value parameterization of genetic values*.
For comparison we also show *genotypic parameterization of genetic values*.
All these quantities can be obtained using `genParam`.
We will return to theory of non-additive deviations after this demonstration.

```{r}
# Observed genotype frequencies
tmp = table(popQtlGenos)
(fObs = prop.table(tmp))

# Allele frequencies and Hardy-Weinberg genotype frequencies
(tmp = mean(popQtlGenos) / 2)
(fHW = c((1 - tmp)^2, 2 * tmp * (1 - tmp), tmp^2))

# Deviation from Hardy-Weinberg equilibrium frequencies
fObs - fHW

# Calculate the quantities
ans = genParam(pop)

# Alpha under
head(ans$alpha[[1]])    # Observed frequencies
head(ans$alpha_HW[[1]]) # Hardy-Weinberg frequencies
```

```{r}
# Quantities from
# genotypic parameterization of genetic values and
# breeding value parameterization of genetic values
tmp = cbind(popQtlGenos, ans$gv[, 1],
            ans$gv_mu[1], ans$gv_a[, 1], ans$gv_d[, 1],
            ans$mu[1], ans$bv[, 1], ans$dd[, 1])
colnames(tmp) = c("genotype", "gv",
                  "mu0", "x_A_a", "x_D_d",
                  "meanGv", "bv", "dd")
head(tmp)
# ... genetic value for the individual i
id = 2 # change to inspect other individuals
tmp[id, "gv"]
# ... and its breeding value parameterization
tmp[id, "meanGv"] + tmp[id, "bv"] + tmp[id, "dd"]

# Mean of breeding values and deviations is (numerical) zero
mean(ans$bv)
mean(ans$dd)
# ... while this is not the case for genotypic parameterization
mean(ans$gv_a)
mean(ans$gv_d)
```

### Visualizing breeding values at one locus

We now have all the ingredients to reproduce the well-known figure of
breeding value parameterization of genetic values from @falconer1996Introduction.
We will mirror AlphaSimR calculation by
showing the projection of genetic values onto dosages.

First, prepare all the ingredients.

```{r}
gvs = ans$gv
# also gv(pop)
meanGv = ans$mu
# also mean(gv(pop))
# also meanG(pop)

dosage = popQtlGenos
meanDosage = mean(dosage)
x = dosage - meanDosage

# alpha from observed genotypes
cov(gvs, x) / var(x)
# also ans$alpha[[1]]

bvs = ans$bv
# also bv(pop)
# also x %*% ans$alpha[[1]]
```

Now the figure.

```{r}
rangeY = range(c(gvs, meanGv + bvs))
# ... genetic values (black dots)
plot(gvs ~ dosage,
     xlab = "Genotype dosage", ylab = "Genetic value",
     ylim = rangeY, pch = 21, bg = "black")
# ... least-squares line
abline(a = meanGv - ans$alpha[[1]] * meanDosage, b = ans$alpha[[1]])
# ... breeding values (white dots on the line)
points(meanGv + bvs ~ dosage, pch = 21, bg = "white")
# ... deviations between genetic values and breeding values
segments(x0 = dosage, x1 = dosage,
         y0 = gvs, y1 = meanGv + bvs)
# ... mean of genetic values and dosages (plus symbol)
points(meanGv ~ mean(dosage), pch = "+")
```

### Extension to multiple loci

It is instructive to demonstrate the same principles with multiple QTL.

```{r}
founderPop = quickHaplo(nInd = 100, nChr = 1, segSites = 3)
SP = SimParam$new(founderPop)
SP$addTraitAD(nQtlPerChr = 3, mean = 10, meanDD = 0.75)
SP$traits
pop = newPop(founderPop)
popQtlGenos = pullQtlGeno(pop)

# Observed genotype frequencies
tmp = apply(popQtlGenos, 2, table)
(fObs = apply(tmp, 2, prop.table))

# Allele frequencies and Hardy-Weinberg genotype frequencies
(tmp = apply(popQtlGenos, 2, mean) / 2)
(fHW = rbind((1 - tmp)^2, 2 * tmp * (1 - tmp), tmp^2))

# Deviation from Hardy-Weinberg equilibrium frequencies
fObs - fHW

# Calculate the quantities
ans = genParam(pop)

# Alphas under
head(ans$alpha[[1]])    # Observed frequencies
head(ans$alpha_HW[[1]]) # Hardy-Weinberg frequencies

# Quantities from
# breeding value parameterization of genetic values and
# genotypic parameterization of genetic values
tmp = cbind(ans$gv[, 1],
            ans$gv_mu[1], ans$gv_a[, 1], ans$gv_d[, 1],
            ans$mu[1], ans$bv[, 1], ans$dd[, 1])
colnames(tmp) = c("gv",
                  "mu0", "x_A_a", "x_D_d",
                  "meanGv", "bv", "dd")
head(tmp)
```

### Dominance deviations

Once the breeding value has absorbed everything that can be explained by
a linear regression on dosage, whatever is left over at that locus is the
non-additive deviation. It is simply defined as a lack of fit.
In the demonstration above the dominance deviation is obtained as:

$$D(x) = G(x) - A(x) - \bar{G},$$

so that $G(x) = \bar{G} + A(x) + D(x)$ exactly.
Because $A(x)$ is a least-squares projection,
the dominance deviations are uncorrelated with
the breeding values within a locus,
and they have mean zero over the population.

These dominance deviations are often referred to as dominance values.
Critically, these deviations/values $D(x)$ are not
the dominance genotypic effects $d$ or their contributions to genetic values $x_D d$!
The same holds for breeding values $A(x)$ - they are not
additive genotypic effects $a$ or their contributions to genetic values $x_A a$!
This is a constant source of confusion [@hill2008Data, @hivert2021Gene].
Breeding values $A(x)$ are a linear projection of genetic values $G(x)$ onto dosages $x$
(capturing additive and partly also non-additive genotypic effects),
while dominance deviations $D(x)$ are residuals of this projection
(capturing the remainder of genetic values when only additive and dominance effects exist).
These residuals are non-zero when there are non-zero non-additive genotypic effects
as these effects make $G(x)$ non-linear in $x$.

### Epistatic deviations

Epistatic deviations $AA(x)$ follow the same lack-of-fit logic,
applied to a pair of loci rather than to a single locus.
The calculation proceeds in two stages.

#### Stage 1: marginal genetic values

For the pair $(i,j)$,
the marginal genetic value of dosage $x_i$ at the locus $i$ is
its genetic value averaged over the distribution of dosages at the locus $j$:

$$G(x_i) = \sum_{x_j} f(x_j) \, G(x_i, x_j),$$

and symmetrically for the locus $j$.
Because the epistatic contribution to genetic value is
$x_{A_i} x_{A_j} e_{(i,j)}$,
the average over the locus $j$ contributes
$x_{A_i} \bar{x}_{A_j}$ e_{(i,j)},
which is linear in $x_{A_i}$.
The epistatic effect therefore shifts
the marginal genetic value at the first locus ($i$) in
a way that depends on the allele frequency at the second locus $(j)$.
Note that this is the mechanism
by which epistasis contributes to breeding values and their variance (additive genetic variance) and
changes in allele frequencies can thereby "convert" epistatic into additive variance [@hill2017Conversion].

The average effect $\alpha_i$ and breeding values $A(x)_i$ for locus $i$
are then obtained from the marginal genetic values exactly as in the
single-locus case with linear regression shown above, and
the dominance deviations $D(x)_i$ are residuals from that regression.
Because the marginalized epistatic contribution is linear in $x_{A_i}$,
it is absorbed entirely by the linear regression for breeding values and
leaves the dominance deviations untouched.
Epistasis in this model generates additive and epistatic variance,
but no dominance variance.

#### Stage 2: epistatic deviations

The two-locus breeding values and dominance deviations are
the sums of the single-locus quantities:

$$A(x_i, x_j) = A(x_i) + A(x_j) \qquad \text{and} \qquad
D(x_i, x_j) = D(x_i) + D(x_j),$$

and the epistatic deviation is what remains:

$$AA(x_i, x_j) = G(x_i, x_j) - A(x_i, x_j) - D(x_i, x_j) - \bar{G}.$$

Summing over the pairs gives the individual's epistatic deviation for the trait.
As with the other components it has mean zero over the population,
so that for every individual the following holds exactly:

$$G(x) = \bar{G} + A(x) + D(x) + AA(x).$$

Let's demonstrate this.

```{r}
founderPop = quickHaplo(nInd = 100, nChr = 2, segSites = 20)
SP = SimParam$new(founderPop)
SP$addTraitADE(nQtlPerChr = 10, mean = 10, meanDD = 0.75, relAA = 0.2)
pop = newPop(founderPop)
ans = genParam(pop)
tmp = cbind(ans$gv[, 1],
            ans$gv_mu[1], ans$gv_a[, 1], ans$gv_d[, 1], ans$gv_aa[, 1],
            ans$mu[1], ans$bv[, 1], ans$dd[, 1], ans$aa[, 1])
colnames(tmp) = c("gv",
                  "mu0", "x_A_a", "x_D_d", "x_AA_aa",
                  "meanGv", "bv", "dd", "aa")
head(tmp)

# G = mu + A + D + AA, for every individual
resid = ans$gv[, 1] - (ans$mu[1] + ans$bv[, 1] + ans$dd[, 1] + ans$aa[, 1])
max(abs(resid)) # numerical zero for every individual
```

#### Interpretation under linkage disequilibrium

One qualification is worth stating.
The marginalization above weights the partner locus by its own marginal genotype frequencies,
which treats the two loci of a pair as independent.
When they are in linkage disequilibrium (LD)
this need not be the least-squares projection in the observed population,
and the components that follow from it are not necessarily orthogonal [@hill2015Expected].
This is why the covariances between the components are reported
rather than assumed to be zero.

The consequence is a difference in
labeling decompositions of genetic variance
rather than a loss of variance.
Under LD,
additive-by-additive effects can induce a marginal dominance component,
which a parameterization built on the joint two-locus genotype frequencies
reports as dominance variance [@alvarezcastro2019Orthogonal].
AlphaSimR's implementation can not accommodate this.
Because the marginalization makes the epistatic contribution exactly linear
in a locus's own dosage,
a trait with no dominance effects has
dominance deviations of exactly zero however strong the disequilibrium, and
the variance in question is reported as additive-by-additive instead.
The genetic values and their variance are however unaffected
because $G(x) = \bar{G} + A(x) + D(x) + AA(x)$ as shown above.
This is one of the reasons why AlphaSimR also reports
the sum of all non-additive deviations $G(x) = \bar{G} + A(x) + N(x)$ as shown below.

Two things limit how much this consequence matters in practice.
AlphaSimR assigns epistatic QTL pairs by a random permutation across the genome,
so interacting pairs are not preferentially tightly linked.
Recombination erodes LD between loosely linked loci,
while drift, selection, and population structure can generate new LD.
Also, in the theoretical genotype distribution of an unselected diploid F2
population derived from two inbred parents,
additive-by-additive contribution is uncorrelated with the additive dosage
predictors, even under LD. This preserves the additive projection,
but does not make the full variance decomposition orthogonal.

### Total non-additive deviations

AlphaSimR also reports the total non-additive deviation as $N = D + AA = G - \bar{G} - A$,
available through `nd(pop)` or `genParam(pop)$nd`.
It has population mean zero because its components have mean zero.
For traits without epistasis it equals the dominance deviation;
without dominance it equals the epistatic deviation; and
for purely additive traits it is zero.

```{r}
tmp = cbind(ans$gv[, 1],
            ans$gv_mu[1], ans$gv_a[, 1], ans$gv_d[, 1], ans$gv_aa[, 1], ans$gv_n[, 1],
            ans$mu[1], ans$bv[, 1], ans$dd[, 1], ans$aa[, 1], ans$nd[, 1])
colnames(tmp) = c("gv",
                  "mu0", "x_A_a", "x_D_d", "x_AA_aa", "x_N_n",
                  "meanGv", "bv", "dd", "aa", "nd")
head(tmp)

# G = mu + A + D + AA = mu + A + N, for every individual
resid = (ans$dd[, 1] + ans$aa[, 1]) - ans$nd[, 1]
max(abs(resid)) # numerical zero for every individual
```

This is a quantity from the breeding value parameterization,
distinct from `genParam(pop)$gv_n`,
the contribution of non-additive genotypic effects to genetic values.

## Genetic and genic variances {#genetic-variances}

### Variances and covariances

Having decomposed genetic values of individuals,
we can also decompose genetic variance of their population.
The genetic variance is defined simply as
variance of genetic values in a population:

$$\sigma^2_G = \frac{\sum_i \big(G_i - \bar{G}\big)^2}{n},$$

where summation is across individuals $(i)$ of a population with size $n$.
Note that AlphaSimR reports population variance, where we divide by $n$,
instead of sample variance, where we divide by $n-1$.
This leads to only a small difference for large populations,
but can be substantial for small populations.

```{r}
# Genetic variance
# = variance of genetic values
varG(pop)
k = (nInd(pop) - 1) / nInd(pop)
# also var(gv(pop)) * k
# also popVar(gv(pop))
# also ans$varG

var(gv(pop)) # sample variance, while AlphaSimR reports population variance!

# Difference between population and sample variance in a small population
nSmall = 10
# ... population variance
popVar(gv(pop[1:nSmall]))
# ... sample variance
var(gv(pop[1:nSmall]))
```

### Three reference populations

Having access to breeding values and non-additive deviations,
we can calculate their variances and covariances in the same way.
AlphaSimR does this for three different "reference" populations
(observed population, linkage-equilibrium population, and idealized population)
and understanding these is the key to understanding the output of `genParam`.

The **observed population** is
the set of individuals in a population and uses their actual genotypes.
Variances computed here are the *genetic* variances of the population
(`varG` for $\sigma^2_G$, `varA` for $\sigma^2_A$, etc.).
These variances include deviations from Hardy-Weinberg or other equilibria,
such as linkage disequilibrium (LD).
This is best demonstrated with AlphaSimR code.

```{r}
# Additive genetic variance
# = variance of additive genetic (breeding) values
varA(pop)
# also var(bv(pop)) * k
# also popVar(bv(pop))
# also ans$varA

# Dominance genetic variance
# = variance of dominance deviations
varD(pop)
# also var(dd(pop)) * k
# also popVar(dd(pop))
# also ans$varD

# Epistatic genetic variance
# = variance of epistatic deviations
varAA(pop)
# also var(aa(pop)) * k
# also popVar(aa(pop))
# also ans$varAA

# Non-additive genetic variance
# = variance of non-additive deviations
varN(pop)
# also var(nd(pop)) * k
# also popVar(nd(pop))
# also ans$varN

# Covariances (we describe these covariance in a separate section below)
# Because G = A + D + ... = A + N then
# varG = varA + varD + 2covAD + ... = varA + varN + 2covAN
(tmp = popVar(cbind(bv(pop), dd(pop), aa(pop), nd(pop))))
# also cov(bv(pop), dd(pop)) * k
# also cov(bv(pop), aa(pop)) * k
# also cov(dd(pop), aa(pop)) * k
# also cov(bv(pop), nd(pop)) * k
varG(pop)
sum(tmp[1:3, 1:3]) # varA + varD + 2covAD + ...
sum(tmp[c(1,4), c(1,4)]) # varA + varN + 2covAN
```

The **linkage-equilibrium population** has
the same genotype frequencies at each locus as the observed population,
but loci are treated as independent (non-correlated).
Variances computed here are sums of single-locus
(or single-pair for epistasis) contributions
with no covariance (LD) between loci.
We will demonstrate the linkage-equilibrium population variances below.

The **idealized population** has
the same *allele* frequencies as the observed population,
but genotype frequencies at Hardy-Weinberg proportions and loci in linkage equilibrium.
These variances are the *genic* variances:
`genicVarG` for $\sigma^{2}_{G,\text{genic}}$,
`genicVarA` for $\sigma^{2}_{A,\text{genic}}$,
etc.
The genic variance is the variance the population would have
at the same allele frequencies under Hardy-Weinberg and linkage equilibrium.
We will demonstrate the idealized population variances below.

The genic variances are computed as
exact theoretical expectations over the frequency distribution
rather than variances over observed individuals.
For the additive component,
let $A_{HW}(x_i) = (x_i - \bar{x}_i)\alpha_{HW,i}$,
then we have:

$$\sigma^{2}_{A,\text{genic}} = \sum_i \sum_x f_{HW}(x_i)\, A_{HW}(x_i)^{2}$$

and similarly for other components.
Because breeding values and deviations have mean zero,
these sums of squares are variances.
The total genic variance is then sum of the component genic variances.

Following the distinction between
Hardy-Weinberg and linkage disequilibrium contributions described by @bulmer1976Effect,
AlphaSimR expresses genetic variance as genic variance plus two additional terms.
For additive genetic variance we have:

$$\sigma^{2}_{A} = \sigma^{2}_{A,\text{genic}} + \sigma_{A,HW} + \sigma_{A,L}.$$

Here, $\sigma^{2}_{A,\text{genic}}$ (`genParam(pop)$genicVarA`)
is the variance in the idealized population.
The first addition, $\sigma_{A,HW}$ (`genParam(pop)$covA_HW`),
is the variance in the linkage-equilibrium population minus that in the idealized population.
It measures the change caused by replacing Hardy-Weinberg genotype frequencies
with observed frequencies while keeping loci independent.
The second addition, $\sigma_{A,L}$ (`genParam(pop)$covA_L`),
is the variance in the observed population minus that in the linkage-equilibrium population.
It accounts for linkage disequilibrium (LD) between loci.
Thus, adding both terms gives the full $\sigma^{2}_{A}$ (`genParam(pop)$varA`).

The same partition applies to the additive, dominance, epistatic,
total non-additive components, and total genetic variance.
Non-random mating, drift, and selection can change these terms
through changes in genotype frequencies and linkage disequilibrium between loci,
even when allele frequencies change little.
We demonstrate departures from Hardy-Weinberg and linkage equilibrium in the following.
For a more comprehensive study see @lara2022Temporal.

### Departures from Hardy-Weinberg proportions {#hardy-weinberg-departures}

`covX_HW` measures the difference in total variance between the
linkage-equilibrium and idealized populations,
it compares the variance at observed genotype frequencies with
that at Hardy-Weinberg genotype frequencies,
using the corresponding substitution effects in each case.
Inbreeding generally makes this term positive
because homozygotes are over-represented.

### Worked example: equal allele frequencies, different inbreeding

We construct two diploid populations with one QTL,
the same genotypic effects,
but different genotype frequencies.
In the first population we have genotype frequencies
$(0.64, 0.32, 0.04)$ for dosages $(0, 1, 2)$,
exactly as expected with Hardy-Weinberg equilibrium (HWE) at $p = 0.2$.
In the second population we have genotype frequencies
$(0.8, 0, 0.2)$ hence fully inbred,
with the same allele frequency.
With only one locus, there is no LD.

```{r}
set.seed(113)
founderDemo = quickHaplo(nInd = 1000, nChr = 1, segSites = 1)
SPDemo = SimParam$new(founderDemo)
SPDemo$addTraitAD(nQtlPerChr = 1, meanDD = 0.5)
popDemo = newPop(founderDemo, simParam = SPDemo)

# One representative of each dosage, repeated to set exact frequencies
representatives = match(0:2, as.vector(pullQtlGeno(popDemo, simParam = SPDemo)))
stopifnot(!anyNA(representatives))
popHWE = popDemo[rep(representatives, times = c(64, 32, 4))]
popInbred = popDemo[rep(representatives, times = c(80, 0, 20))]

referenceSummary = function(population) {
  g = genParam(population, simParam = SPDemo)
  c(p = mean(pullQtlGeno(population, simParam = SPDemo)) / 2,
    alpha = g$alpha[[1]][1],
    alpha_HW = g$alpha_HW[[1]][1],
    genicVarG = unname(g$genicVarG[1]),
    HW_adjustment = unname(g$covG_HW[1]),
    observed_frequency_VarG = unname(g$genicVarG[1] + g$covG_HW[1]),
    observed_VarG = g$varG[1, 1],
    LD_adjustment = unname(g$covG_L[1]),
    genicVarD = unname(g$genicVarD[1]),
    observed_VarD = g$varD[1, 1])
}
round(rbind(HWE = referenceSummary(popHWE),
            inbred = referenceSummary(popInbred)), 4)
```

The allele frequency, `alpha_HW`, and genic variances
are identical in the two rows.
However, `alpha` changes: under Hardy-Weinberg equilibrium it is
$\alpha_{HW} = a + (1 - 2p)d = a + 0.6d$,
whereas in the fully inbred population it is $\alpha = a$,
because only the two homozygotes remain.
Their genetic values have not changed;
the regression defining the breeding values has changed.

In the Hardy-Weinberg equilibrium row, `HW_adjustment` is zero.
In the inbred row,
`observed_frequency_VarG` includes the adjustment and equals `observed_VarG`.
`LD_adjustment` is zero in both rows since we are working with one QTL.
The inbred population also has zero observed dominance variance,
despite non-zero dominance genotypic effect and positive genic dominance variance:
its Hardy-Weinberg equilibrium reference includes heterozygotes,
while the observed population contains none.

For a purely additive one-locus trait,
full inbreeding at fixed allele frequency doubles variance,
from $2p(1-p)a^2$ to $4p(1-p)a^2$.
With dominance or epistasis,
a doubling of total genetic variance is not a general expectation.

### Linkage disequilibrium and the Bulmer effect {#disequilibrium}

After accounting for genotype frequencies at each locus,
the remaining additional term $\sigma_{A,L}$ (`genParam(pop)$covA_L`),
measures the difference between the observed and linkage-equilibrium populations.
For additive variance,
`covA_L` is the covariance contribution between loci arising from LD.
It is obtained by subtracting `genicVarA` and `covA_HW` from
the observed additive variance `varA`.

The change in additive genetic variance caused by selection-induced LD
is known as the Bulmer effect [-@bulmer1971Effect].
Directional or stabilizing selection can generate a negative LD contribution,
whereas disruptive selection can generate a positive contribution.
Recombination reduces these associations and
can therefore reveal some of the variance masked by negative LD.

### Worked example: selection changes genetic and genic variance

The Bulmer partitioning is easiest to see by performing selection and
then inspecting what changed.
In the example below the genic additive variance is
almost unchanged after one generation of strong selection,
because allele frequencies have barely moved,
while the additive genetic variance has
dropped because selection has generated negative linkage disequilibrium.

```{r}
set.seed(1234)
founderSelection = quickHaplo(nInd = 1000, nChr = 10, segSites = 20)
SPSelection = SimParam$new(founderSelection)
SPSelection$addTraitA(nQtlPerChr = 10)
SPSelection$setVarE(h2 = 0.5)
popSelection = newPop(founderSelection, simParam = SPSelection)

selectedPop = selectInd(popSelection, nInd = 100, simParam = SPSelection)
selectionProgeny = randCross(selectedPop, nCrosses = 1000, simParam = SPSelection)

beforeSelection = genParam(popSelection, simParam = SPSelection)
afterSelection = genParam(selectionProgeny, simParam = SPSelection)

round(c(varA_before = beforeSelection$varA[1,1],
        varA_after = afterSelection$varA[1,1],
        genicA_before = unname(beforeSelection$genicVarA[1]),
        genicA_after = unname(afterSelection$genicVarA[1]),
        covA_L_before = unname(beforeSelection$covA_L[1]),
        covA_L_after = unname(afterSelection$covA_L[1])), 3)
```

### Choosing a reference population {#choosing-reference-population}

The two examples illustrate why the choice of reference population matters.
In most cases,
users will be interested in variances in the observed population.
The other two reference populations are useful
when we try to understand why variances differ.
For `ans = genParam(pop)`, we have:

| Population | Genotype frequencies | Association between loci (LD) | Total variance |
|:-----------|:---------------------|:------------------------------|:---------------|
| Observed | Observed | Observed | `varG(pop)` or `ans$varG` |
| Linkage equilibrium | Observed | Independent | `ans$genicVarG + ans$covG_HW` |
| Idealized | HWE at current allele frequencies | Independent | `ans$genicVarG` |

Thus, to retain departures from Hardy-Weinberg equilibrium but removing LD,
use `ans$genicVarG + ans$covG_HW`.
This is generally **not** the variance of genetic values in the observed population:
recovering that variance also requires `ans$covG_L`.
The same recipe applies with `G` replaced by `A`, `D`, `AA`, or `N`.
Note that changing reference population also changes the substitution effects:
the idealized calculation uses `alpha_HW`,
while the observed-frequency calculation uses `alpha`.
Adding `covX_HW` therefore accounts for changes in
both genotype frequencies and substitution effects.

### Covariances between genetic components {#component-covariances}

Returning to the population with dominance and epistasis used earlier,
we can also examine covariances between its genetic components.
Because the total genetic value is a sum of multiple components,
the total genetic variance also includes covariances *between* these components.
These are reported separately as `covAD_L`, `covAAA_L`, `covDAA_L`, and `covAN_L`,
respectively the covariances
between breeding values and dominance deviations,
between breeding values and epistatic deviations,
between dominance and epistatic deviations, and
between breeding values and non-additive deviations.
All these are zero in the idealized population and non-zero in general.
The complete identity is:

$$
\begin{aligned}
\sigma^{2}_{G}
& = \sigma^{2}_{A} + \sigma^{2}_{D} + \sigma^{2}_{AA}
    + 2\,\sigma_{A\cdot D,L}
    + 2\,\sigma_{A\cdot AA,L}
    + 2\,\sigma_{D\cdot AA,L}, \\
& = \sigma^{2}_{A} + \sigma^{2}_{N}
    + 2\,\sigma_{A\cdot N,L},
\end{aligned}
$$

which follows directly from $G = \bar{G} + A + D + AA = \bar{G} + A + N$ and evaluating $Var(G)$.

```{r}
# The full variance identity
varG = ans$varG[1,1]
varG_via_partsADAA = unname(ans$varA[1,1] + ans$varD[1,1] + ans$varAA[1,1] +
  2*(ans$covAD_L[1] + ans$covAAA_L[1] + ans$covDAA_L[1]))
varG_via_partsAN = unname(ans$varA[1,1] + ans$varN[1,1] +
  2*(ans$covAN_L[1]))

c(varG = varG, varG_via_partsADAA = varG_via_partsADAA, varG_via_partsAN = varG_via_partsAN)
```

## Polyploids and digenic dominance {#polyploids}

### Genotype coding and model limitations

AlphaSimR uses one dominance genotypic effect $d$ per locus.
For diploids, this describes the departure of the heterozygote's genetic value
from the midpoint of the two homozygotes.
An autotetraploid locus has three heterozygous genotypes,
with dosages 1, 2, and 3.
Describing their departures independently requires three dominance parameters,
conventionally expressed as digenic, trigenic, and quadrigenic effects,
respectively involving two, three, and four alleles at the locus [@gallais2003Quantitative].

AlphaSimR models only digenic dominance.
Its scaled dominance dosage, $x_D = x(P-x)(2/P)^2$,
is proportional to the number of heterozygous allele pairs:
each of the $x$ copies of the "1" allele pairs with
each of the $P-x$ copies of the "0" allele.
The dominance contribution $x_D d$ therefore follows a fixed pattern across
genotypes rather than allowing a separate effect for each heterozygote.
The scaled dosages $x_A$ and $x_D$ for diploids and autotetraploids are
illustrated in `vignette("traits")`.

This restriction concerns the assumed genotypic model,
not the completeness of its decomposition.
The identity $G = \bar{G} + A + D + AA$ still holds for polyploids,
with dominance deviations calculated from the modeled genetic values.
AlphaSimR does not represent trigenic or quadrigenic dominance genotypic effects,
so its reported dominance variance reflects only the digenic dominance
included in the model.

### Hardy-Weinberg reference and double reduction

There is a second polyploid-specific caveat.
This one concerns the Hardy-Weinberg reference.
The binomial genotype frequencies used to
compute the genic variances are the correct polysomic equilibrium only
under random chromosome segregation without double reduction.
Quadrivalent pairing,
described in the `vignette("GeneticRecombination")`,
produces double reduction and therefore an excess of homozygotes
relative to the binomial expectation.
AlphaSimR does not model this in its Hardy-Weinberg reference,
so in a simulation with `SP$quadProb > 0`
the resulting excess homozygosity appears in the `covD_HW` and `covA_HW` terms
as though it were non-random mating.
This is arguably the right place for it,
since the effect on the variance is the same,
but it means those terms should not be read
as a pure measure of the mating system in an autopolyploid simulation.

## Relationship to classical theory {#classical-theory}

In the idealized population, the decomposition described here follows the
least-squares partition introduced by @fisher1919Correlation and extended
to epistasis by @cockerham1954Extension and @kempthorne1954Correlation.
The additive-by-additive contrast used by AlphaSimR
is the product of the two single-locus additive contrasts,
which is the standard construction and the one that makes the components
orthogonal in an idealized population.
In the terminology of the natural and orthogonal interactions framework of @alvarezcastro2007Unified,
AlphaSimR computes the breeding value parameterization
against the observed genotype frequencies,
which is what makes `alpha` and `alpha_HW` differ.

Orthogonality of the components holds in the idealized population,
with Hardy-Weinberg proportions and linkage equilibrium.
In an observed population the components can be correlated.
The cross-component covariances are reported explicitly as
`covAD_L`, `covAAA_L`, and `covDAA_L`, rather than being assumed zero.
Consequently, recovering `varG` from `varA`, `varD`, and `varAA` generally
requires the cross-component covariance terms.

Two conventions are available for a population away from equilibrium,
and it is worth knowing which one is in use.
The first convention defines the effects in an equilibrium reference population and
lets disequilibrium appear as covariances between components,
leaving the effects themselves unchanged;
this is the convention of @wang2006Models,
which makes the Bulmer partitioning above interpretable.
The second convention refits the effects against the observed multilocus genotype frequencies,
which restores orthogonality at the cost of absorbing the disequilibrium into
the components themselves;
@alvarezcastro2019Orthogonal give such a decomposition for epistatic traits.
AlphaSimR retains covariances between components while recomputing the
average effects of allele substitution from the current marginal genotype frequencies.

One further point of interpretation, made forcefully by @hill2008Data,
is worth keeping in mind when setting up a simulation with epistasis.
A model with substantial additive-by-additive epistasis genotypic effects
will usually produce a population in which most of the genetic variance is
captured by statistical additive genetic variance,
particularly when allele frequencies are away from one half.
The `relAA` argument specifies the ratio of epistatic to additive variance
expected in a population at allele frequency 0.5 in linkage equilibrium;
observed ratios in a simulated population will typically be smaller,
sometimes much smaller.
This is not an artifact of the implementation
but a property of the decomposition itself.

Lastly,
much of the classical quantitative genetics theory
concerns covariances between relatives and
the partitioning of phenotypic and genetic variance within and between families.
Readers can explore these relationships by combining
the genetic values and variance calculations introduced here
with `parentAverage`, `mendelianSampling`, and simulated crosses.
The companion `vignette("GeneticRecombination")`` explains
the recombination mechanisms underlying those crosses.
As an exercise, simulate several families and compare variation in
parental averages with variation among offspring within each family.

## References
