---
title: "Tree-sequence export"
author: "AlphaSimR contributors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tree-sequence export}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

AlphaSimR can export its recorded pedigree and recombination history as
[tskit](https://tskit.dev/) tree sequences. Export is deliberately on demand:
AlphaSimR continues to use its native genotype representation while running a
breeding-program simulation, and creates tree sequences only when requested.
This preserves the behavior and performance assumptions of existing AlphaSimR
scripts.

## Recording and exporting ancestry

Recombination tracking must be enabled before the first population is created.

```{r eval=FALSE}
library(AlphaSimR)

founderPop = quickHaplo(nInd=20, nChr=3, segSites=100)
SP = SimParam$new(founderPop)
SP$setTrackRec(TRUE)

founders = newPop(founderPop, simParam=SP)
generation1 = randCross(founders, nCrosses=20, simParam=SP)
generation2 = randCross(generation1, nCrosses=20, simParam=SP)

trees = asTreeSequence(generation2, simParam=SP)
writeTreeSequence(trees, "breeding-program.trees")
```

One tree sequence is returned per requested chromosome. When multiple
chromosomes are written, the output names receive `_chr1`, `_chr2`, and so on.
By default, each tree sequence is simplified to the requested population while
retaining the founder-haplotype roots. Set `simplify=FALSE` when the complete
recorded pedigree is needed.

The returned objects can be inspected directly in R. For example, the first
chromosome reports its sample, tree, site, and mutation counts through the
RcppTskit interface.

```{r eval=FALSE}
trees[[1]]$num_samples()
trees[[1]]$num_trees()
trees[[1]]$num_sites()
trees[[1]]$num_mutations()
```

Files written by `writeTreeSequence()` use the standard tskit `.trees` format
and can be opened in Python without AlphaSimR.

```{python eval=FALSE}
import tskit

ts = tskit.load("breeding-program_chr1.trees")
print(ts)
genotypes = ts.genotype_matrix()
print(ts.metadata)
```

When a list containing more than one chromosome is written to
`"breeding-program.trees"`, AlphaSimR creates
`breeding-program_chr1.trees`, `breeding-program_chr2.trees`, and so on. A
single requested chromosome is written to the path supplied by the user.

## Samples and variants

The samples are ordered exactly as the individuals and homologs in the supplied
`Pop`. A `MultiPop` can be used to sample populations from different
generations or with different ploidies, provided their chromosome and locus
structures agree. The same recorded individual (identified internally by its
AlphaSimR `iid`) cannot be supplied twice. Public `Pop@id` values are metadata,
need not be unique, and are written exactly as they appear in the sampled
population. This includes IDs assigned by `pedigreeCross()` or changed by the
user after a population was created.

With `includeVariants=TRUE`, every simulated locus is exported and the current
sample haplotypes are reproduced exactly. Founder alleles are placed on shared
founder-origin nodes when sampled tracked founders verify the correspondence
between recorded origins and `SimParam$founderPop`. Otherwise they are inferred
conservatively from the requested samples. Alleles introduced by `mutate()`,
`editGenome()`, an additional founder, or a reordered founder population are
represented parsimoniously when the recorded data allow it; remaining
differences are represented as mutations on the affected sample lineages.
These terminal sample lineages also keep a post-hoc edit to a sampled ancestor
from changing descendants that were generated before the edit. This means
current sample genotypes are exact even when AlphaSimR does not retain enough
information to locate the historical event that produced a difference.

Set `includeVariants=FALSE` for ancestry-only output. This avoids constructing
the allele matrices and is preferable when only pedigree/recombination topology
is required.

## Coordinates and metadata

Tree-sequence coordinates are locus indices rather than physical base-pair
positions. A chromosome with `n` loci spans `[0, n)`, and locus `i` is at
position `i - 0.5`. Recombination breakpoints lie on integer boundaries between
loci. The coordinate system, chromosome, AlphaSimR version, individual IDs,
homolog numbers, and locus indices are recorded as JSON metadata. Provenance
records identify `asTreeSequence()` as the operation that created the file.
The tables carry permissive JSON metadata schemas, so current Python tskit
versions return decoded dictionaries from `ts.metadata`,
`individual.metadata`, `node.metadata`, and `site.metadata`.

Sample and recorded-individual node times are pedigree depths in generations,
measured backwards from the deepest individual in the complete recorded
pedigree. That deepest generation has time zero. Consequently, a sampled
ancestor or a sample from a shallower branch can have a positive time, even if
no descendant of that sample is also exported. Synthetic founder-origin roots
are one generation older than founder lineage nodes. A sampled individual's
internal ancestry proxy is placed at the next representable older floating-
point time so that a current-genome override affects that sample without
changing existing descendants. The top-level metadata records the origin,
scale, founder-root offset, and proxy convention explicitly.

## Representation limits

`HybridPop` cannot be exported because it intentionally stores genetic values
rather than haplotypes and recombination histories. Use
`hybridCross(..., returnHybridPop=FALSE)` when tree-sequence output is needed.

Histories generated by AlphaSimR versions containing the old quadrivalent
homolog-remapping bug cannot be reconstructed reliably. Export detects those
records and asks that the simulation be regenerated. Newly generated
quadrivalent, doubled, reduced, merged, doubled-haploid, selfed, and
unequal-ploidy histories are supported.

The `simParam` argument must be the same `SimParam` object that recorded the
population. In ancestry-only mode AlphaSimR cannot prove that a different
simulation with coincident internal IDs was not supplied.
