---
title: "Gene Drop Simulations"
author: "Chris Gaynor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gene Drop Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(AlphaSimR)
set.seed(3453)
```

# Introduction

A gene drop simulation takes a pedigree and a set of founder genomes and
works forwards through the pedigree, sampling gametes according to
Mendelian inheritance and recombination, until every individual has a
genome. The pedigree supplies the structure and the founders supply the
genetic material. Nothing about the genomes of the non-founders is
assumed; they are the outcome of the sampling.

Gene drops are useful whenever the question is about how genetic material
moves through a known set of relationships. Typical uses are estimating
the probability that a descendant carries a founder allele, quantifying
how variable realised relationships are around their pedigree
expectations, generating genotypes for individuals that were never
genotyped, and producing a set of starting genomes for a forward
simulation that begins from a real breeding population.

AlphaSimR performs gene drops with `pedigreeCross()`. This article works
through the function in detail, with an emphasis on the common case where
both the pedigree and the founder genotypes come from outside AlphaSimR.

```{r}
library(AlphaSimR)
```

# The pedigree

A pedigree is given to `pedigreeCross()` as three vectors of equal length:
`id`, `mother` and `father`.

These are always character vectors. Anything else is coerced with
`as.character()` before it is used, so an integer pedigree works, but it
becomes character internally and the IDs on the returned population will
be character.

An unknown parent is `NA` by default. This matters, because many pedigree
files use `0` to mean unknown, and AlphaSimR does not read it that way on
its own: a parent recorded as `"0"` names an individual called "0". The
`unknownParent` argument says which values mean unknown, so a file using
`0` is handled by passing `unknownParent = "0"`, and a merged file using
several conventions by passing them all. `NA` is always unknown, whatever
else is given, because it cannot name anything. An `NA` in `id` is an
error, since every individual needs a name, and an `id` that is one of the
`unknownParent` values is an error for the same reason.

The pedigree is extended backwards before it is used, so that every parent
is either another row of the pedigree or unknown. Which names get a row of
their own depends on whether they carry any information about how
individuals are related.

A name used as a parent twice or more, counting the mother and father
columns together, is added as a founder row. Leaving it out would change
the pedigree: two half sibs sharing an unrecorded mother would come back
unrelated, and an individual whose mother and father are the same
unrecorded name would come back as an outcross rather than a self.

A name used once is different. It links its child to nothing else in the
pedigree, so the child can simply be a founder in its own right. Such a
name is treated as unknown and no row is added for it. This pedigree:

```{r}
ped = data.frame(id = "3",
                 mother = "2",
                 father = "1")
```

is therefore a single founder called "3". Neither "1" nor "2" appears
anywhere else, so neither is worth simulating, and individual "3" takes a
genome from `founderPop` directly. If you want the cross, write the
founders out:

```{r}
ped = data.frame(id = c("1", "2", "3"),
                 mother = c(NA, NA, "2"),
                 father = c(NA, NA, "1"))
```

The returned `mother` and `father` show what was actually simulated, so a
name that was dropped this way comes back as `NA` rather than as you wrote
it. Matching on ID is the exception to all of this: with `matchID = TRUE`
every missing name gets a row, because a name has to have a row before it
can be matched against `founderPop`.

The pedigree does not have to be sorted. `pedigreeCross()` sorts it, and in
doing so it will detect a few problems: duplicated IDs, and loops in which
an individual ends up being its own ancestor. The `maxCycle` argument bounds
how many passes over the pedigree are made while sorting; left at `NULL` it
follows the size of the pedigree, which is always enough, so it is not
something you normally need to set.

# A first example

The simplest case is a pedigree dropped onto founders that were simulated
in AlphaSimR. Here is a biparental cross taken to the F8 by single seed
descent.

```{r}
founderGenomes = quickHaplo(nInd = 2, nChr = 2, segSites = 100)
SP = SimParam$new(founderGenomes)
pop = newPop(founderGenomes)

id = as.character(1:10)
mother = c(NA, NA, "1", as.character(3:9))
father = c(NA, NA, "2", as.character(3:9))

f8 = pedigreeCross(founderPop = pop, id = id, mother = mother,
                   father = father)
f8
```

Individuals 1 and 2 have unknown parents, so they are the founders and
they take their genomes from `pop`. Individual 3 is their cross, and
individuals 4 to 10 are successive self-pollinations. The pedigree is
returned with the population.

```{r}
tail(getPed(f8))
```

Because the founders were assigned at random from `pop`, repeating the
call gives a different, equally valid, realisation. That is the point of a
gene drop.

# Working with an external pedigree

Real pedigrees usually arrive as a file. The two things to get right are
the type of the columns and the coding of unknown parents.

`read.csv()` will happily turn an ID column of numbers into an integer
column, and then `"007"` becomes `7`. Reading the identifier columns as
character avoids the problem entirely:

```{r, eval = FALSE}
ped = read.csv("pedigree.csv",
               colClasses = c(id = "character",
                              mother = "character",
                              father = "character"))
```

For this article we will build the same object directly, in the form it
would have after being read from a file that uses `0` for unknown parents.

```{r}
ped = data.frame(
  id     = c("F1", "F2", "F3", "F4", "C1", "C2", "C3", "D1", "D2", "D3"),
  mother = c("0",  "0",  "0",  "0",  "F1", "F3", "C1", "C3", "C3", "C3"),
  father = c("0",  "0",  "0",  "0",  "F2", "F4", "C2", "C3", "C3", "C3")
)
ped
```

Four inbred founders are crossed in pairs, the two single crosses are
crossed to give `C3`, and three doubled haploid lines are derived from
`C3`.

The `0` codes are handled by `unknownParent`, so the file can be used as
it is:

```{r, eval = FALSE}
pedigreeCross(founderPop, ped$id, ped$mother, ped$father,
              unknownParent = "0")
```

A merged pedigree often uses several conventions at once, and they can all
be given together:

```{r, eval = FALSE}
pedigreeCross(founderPop, ped$id, ped$mother, ped$father,
              unknownParent = c("0", "", "NA", "*", "-9"))
```

For the rest of this article we will recode once instead, so that the
calls that follow stay short:

```{r}
ped$mother[ped$mother == "0"] = NA
ped$father[ped$father == "0"] = NA
ped
```

It is worth checking, before simulating, that every named parent is either
a row of the pedigree or unknown. Anything else is a typo, and a typo is
silent: a misspelt parent name used once is simply treated as unknown, and
its child becomes a founder.

```{r}
named = unique(c(ped$mother, ped$father))
named = named[!is.na(named)]
setdiff(named, ped$id)
```

An empty result means the pedigree is complete. Here it is, so nothing is
added and nothing is dropped.

# Working with external genotypes

Founder genomes can come from real data. AlphaSimR reads them with
`importGenMap()` together with either `importInbredGeno()`, for inbred
diploid lines, or `importHaplo()`, for phased haplotypes. Both return a
`MapPop`, or a `NamedMapPop` when a pedigree is supplied with them.

To keep the article self contained, we will manufacture a genotype matrix
and a genetic map of the kind these functions expect. In practice these
two objects would be read from files.

```{r}
# Stands in for real data
sourceGenomes = runMacs(nInd = 6, nChr = 2, segSites = 100, inbred = TRUE)

genoFile = pullSegSiteGeno(sourceGenomes)
rownames(genoFile) = c("F1", "F2", "F3", "F4", "C1", "C2")

mapFile = getGenMap(sourceGenomes)

genoFile[, 1:8]
head(mapFile)
```

The genotype matrix has one row per individual and one column per marker,
coded as allele dosages. The map has three columns: marker name,
chromosome, and position in Morgans. Marker names link the two, so the
column names of the genotype matrix must appear in the map.

```{r, eval = FALSE}
genoFile = as.matrix(read.csv("genotypes.csv", row.names = 1))
mapFile = read.csv("map.csv")
```

Importing gives a founder population:

```{r}
founderPop = importInbredGeno(geno = genoFile,
                              genMap = mapFile,
                              ped = rownames(genoFile))
founderPop
```

Supplying `ped` is what makes the result a `NamedMapPop` rather than a
plain `MapPop`, and it is what allows the founders to be matched to the
pedigree by name in the next section.

## Map populations and SimParam

`pedigreeCross()` treats a map population differently from a population,
because the two say different things about where you are.

Passing a `Pop`, as in the first example, means a simulation is already
running. The pedigree is built with your `SimParam` and the result belongs
to that simulation.

Passing a `MapPop` or `NamedMapPop` means no simulation has been set up
yet, which is the usual situation when founder genotypes have just been
imported. No `SimParam` is looked for, and none is created for you.
`pedigreeCross()` makes a temporary one of its own, uses it to run the
crossing, and discards it. The result is a `NamedMapPop` carrying the
genetic map and the pedigree's IDs, which is exactly what `SimParam$new()`
wants. Because the temporary `SimParam` defines no traits, nothing about
traits or SNP chips survives the call; those are added afterwards, to the
simulation you start from the result.

This gives the natural order of operations for a simulation that begins
from a real population: import the genotypes, drop the pedigree, then set
up the simulation from the dropped population.

# Matching founders by ID

`matchID` decides how individuals in the pedigree are connected to
individuals in `founderPop`.

## matchID = FALSE

With `matchID = FALSE`, the two sets of IDs have nothing to do with each
other. The founders are the individuals whose parents are both unknown,
and each one is assigned a randomly sampled individual from `founderPop`.
Every individual in the extended pedigree is returned.

```{r}
dropped = pedigreeCross(founderPop = founderPop,
                        id = ped$id, mother = ped$mother,
                        father = ped$father,
                        matchID = FALSE)
dropped
getPed(dropped)
```

`founderPop` needs at least as many individuals as the pedigree has
founders. Here there are four founders and six imported lines, so there is
room to spare.

An individual with exactly one unknown parent is a half founder, and that
unknown parent is given a founder genome of its own. A pedigree with
several half founders therefore needs one extra founder for each of them,
and two half founders never share a genome.

```{r}
halfPed = data.frame(id     = c("A", "B", "C", "D"),
                     mother = c(NA,  NA,  NA,  "B"),
                     father = c(NA,  "A", "A", "C"))

halfDrop = pedigreeCross(founderPop = founderPop,
                         id = halfPed$id, mother = halfPed$mother,
                         father = halfPed$father)
halfDrop@id
```

`A` has both parents unknown, so it is a founder. `B` and `C` each have one
unknown parent, and each of those gets a founder genome of its own, so three
founder genomes are drawn from `founderPop` even though the pedigree only
names one individual with two unknown parents. The unknown mothers of `B`
and `C` are not the same individual, and they are not rows of the returned
population; only the named individuals are returned.

The other thing that shapes the founder count is the extension rule from
the first section. A parent named only once is treated as unknown, so an
individual whose parents are both unrecorded singletons is one founder
rather than the cross of two:

```{r}
oncePed = data.frame(id     = c("X", "Y", "Z"),
                     mother = c("m1", "shared", "shared"),
                     father = c("f1", NA, NA))

onceDrop = pedigreeCross(founderPop = founderPop,
                         id = oncePed$id, mother = oncePed$mother,
                         father = oncePed$father)
getPed(onceDrop)
```

`m1` and `f1` are each used once, so both are treated as unknown and `X`
is a founder. `shared` is used twice, so it is added, and `Y` and `Z` come
back as the half sibs the pedigree says they are. Four founder genomes are
used in all: one for `X`, one for `shared`, and one each for the unknown
fathers of `Y` and `Z`.

This keeps the founder count close to what the pedigree actually
constrains, but it is a genetic difference and not only a bookkeeping one.
`X` is now a single individual taken from `founderPop`, where it would
otherwise have been the cross of two. With an inbred `founderPop`, that
means `X` is inbred where it used to be heterozygous.

## matchID = TRUE

With `matchID = TRUE`, an individual whose ID is found in `founderPop`
takes its genotype from there, and its own ancestry is not simulated. Only
the matched individuals and what descends from them are simulated. This is
the setting to use when the founder genotypes are real data for named
individuals.

Because ancestry above a match is skipped, the returned population can be
shorter than the extended pedigree. Our imported population contains `C1`
and `C2` as well as `F1` to `F4`, so matching on ID uses the observed
genotypes of the single crosses and does not simulate the four founders at
all:

```{r}
matched = pedigreeCross(founderPop = founderPop,
                        id = ped$id, mother = ped$mother,
                        father = ped$father,
                        matchID = TRUE)
matched@id
```

All six matched individuals are returned, plus `C3` and the three doubled
haploids that descend from them. Nothing was dropped through `F1` to `F4`,
because their genotypes were observed rather than inferred.

The more common case is that only part of the pedigree is genotyped.
Suppose only the two single crosses were:

Subsetting a map population uses positions rather than names, so
`match()` is the convenient way to pick individuals out of it by ID.

```{r}
genotyped = founderPop[match(c("C1", "C2"), founderPop@id)]

partial = pedigreeCross(founderPop = genotyped,
                        id = ped$id, mother = ped$mother,
                        father = ped$father,
                        matchID = TRUE)
partial@id
```

`F1` to `F4` are legitimately skipped: they are ancestors of a match, so
their genomes are not needed. What is returned is the two matched lines
and their four descendants.

An individual that is neither matched nor descended from a match cannot be
made at all. Rather than inventing a genome for it, `pedigreeCross()`
stops and names the individuals it could not build:

```{r, error = TRUE}
pedigreeCross(founderPop = founderPop[match(c("F1", "F2"), founderPop@id)],
              id = ped$id, mother = ped$mother, father = ped$father,
              matchID = TRUE)
```

`C1` is the cross of two matched founders and could be built, but `C2`
needs `F3` and `F4`, which were not supplied, so the pedigree cannot be
completed.

Matching needs a population that has IDs. A plain `MapPop` does not, so it
can only be used with `matchID = FALSE`. Use `importInbredGeno()` or
`importHaplo()` with their `ped` argument, as above, to get a
`NamedMapPop` instead.

# Starting a simulation from the result

The population returned from a map population is a `NamedMapPop`, so it
can be passed straight to `SimParam$new()`:

```{r}
SP = SimParam$new(matched)
SP$addTraitA(nQtlPerChr = 10)
SP$setVarE(h2 = 0.3)

pop = newPop(matched)
pop
head(getPed(pop))
```

The IDs, mothers and fathers from the pedigree are carried onto the
population, so the simulated individuals keep the names they had in the
original data. From here the simulation continues as any other AlphaSimR
simulation would.

# Doubled haploids and selfing

The three lines `D1` to `D3` in our pedigree are doubled haploids, which
the pedigree alone cannot express: as far as it is concerned they are
three self-pollinations of `C3`. The `DH` and `nSelf` arguments supply the
missing information.

`DH` is a logical vector saying which individuals are doubled haploids.
`nSelf` is an integer vector saying how many generations of selfing an
individual represents, which is useful when a pedigree records a line by
its single cross and its inbreeding is implied.

Both are aligned with the `id` vector as you supply it. Rows added by the
backwards extension are given `DH = FALSE` and `nSelf = 0`, so you never
have to leave space for them.

```{r}
dh = ped$id %in% c("D1", "D2", "D3")

dhDrop = pedigreeCross(founderPop = founderPop,
                       id = ped$id, mother = ped$mother,
                       father = ped$father,
                       matchID = TRUE,
                       DH = dh)
```

Doubled haploids are completely homozygous, which is easy to confirm:

```{r}
geno = pullSegSiteGeno(dhDrop)
any(geno[c("D1", "D2", "D3"), ] == 1)
any(geno["C3", ] == 1)
```

`nSelf` works the same way. An individual recorded as the cross of two
parents, but understood to be an F5 line, is written as the cross with
four generations of selfing after it:

```{r}
selfPed = data.frame(id     = c("C1", "C2", "L1"),
                     mother = c(NA,   NA,   "C1"),
                     father = c(NA,   NA,   "C2"))

selfDrop = pedigreeCross(founderPop = founderPop,
                         id = selfPed$id, mother = selfPed$mother,
                         father = selfPed$father,
                         matchID = TRUE,
                         nSelf = c(0, 0, 4))

mean(pullSegSiteGeno(selfDrop)["L1", ] == 1)
```

The result is a line that is mostly, but not entirely, homozygous, as an
F5 should be.

# Recombination settings

When `founderPop` is a map population, `pedigreeCross()` builds its own
`SimParam`, so the recombination settings it uses can be given through
`...`. The three that are accepted are `v`, the interference parameter,
`p`, the proportion of recombination following a Poisson process, and
`quadProb`, the probability of quadrivalent pairing in autopolyploids. Any
that are not given keep the `SimParam` defaults.

```{r}
noInterference = pedigreeCross(founderPop = founderPop,
                               id = ped$id, mother = ped$mother,
                               father = ped$father,
                               matchID = TRUE,
                               v = 1)
```

These can only be set this way when `founderPop` is a map population.
`SimParam` is an R6 object, with reference semantics, so setting them
during a call that uses your own `SimParam` would change your simulation
permanently rather than just for the call. Passing them with a `Pop` is
therefore an error, and the settings should be made on the `SimParam`
itself:

```{r, error = TRUE}
pedigreeCross(founderPop = pop, id = ped$id, mother = ped$mother,
              father = ped$father, v = 1)
```

# Repeating the drop

A single gene drop is one realisation of a random process. Most questions
about a pedigree are answered by repeating it and looking at the
distribution of the result.

Here we drop the pedigree many times and record the genotype of `C3` at a
single marker. `C1` and `C2` are matched to their observed genotypes every
time, so all of the variation comes from which alleles they transmitted.

```{r}
nRep = 200
dose = numeric(nRep)

for(i in seq_len(nRep)){
  drop = pedigreeCross(founderPop = founderPop,
                       id = ped$id, mother = ped$mother,
                       father = ped$father,
                       matchID = TRUE)
  dose[i] = pullSegSiteGeno(drop)["C3", "1_50"]
}

table(dose)
```

The same approach answers questions about founder contributions, about how
much realised relationships vary around their pedigree expectations, and
about the probability that a particular descendant carries a particular
founder allele. Because the drops are independent, the loop parallelises
trivially.

The drops are also reproducible. Setting the seed before the loop fixes
the whole sequence, and this is worth doing whenever a result is going to
be reported.

# Things to watch for

**Unknown parents must be `NA`.** A `0` in a parent column names an
individual called "0", and because every row with that parent then shares
a single genome, unrelated individuals come out related. Pass
`unknownParent = "0"`, or recode the file before dropping it.

**A parent named once is not simulated.** Under `matchID = FALSE`, a
parent that appears nowhere else in the pedigree is treated as unknown and
its child becomes a founder. This is usually what you want, and it keeps
the founder count down, but it means a misspelt parent name fails quietly.
The `setdiff()` check above catches it.

**`matchID = FALSE` samples founders at random.** The IDs in the pedigree
and the IDs in `founderPop` are unconnected under this setting, so a
pedigree whose founders happen to share names with genotyped individuals
will not use those genotypes. Use `matchID = TRUE` when the names are
meant to line up.

**Separate sexes are not supported.** `pedigreeCross()` currently requires
a `SimParam` with `sexes = "no"`. The pedigree itself determines who
crosses with whom, so the distinction is not needed, but an existing
simulation that uses separate sexes cannot be used as the founder
population.

**Traits do not survive the map population route.** The temporary
`SimParam` defines no traits, so add traits to the `SimParam` you build
from the returned population, not before.

**Check the size of the result.** With `matchID = TRUE` the returned
population is the matched individuals and their descendants, which is
often smaller than the pedigree supplied. Comparing `nInd` against the
pedigree, or inspecting `getPed()` on the result, is a quick way to
confirm that the drop did what was intended.

# Summary

`pedigreeCross()` drops founder genomes through a pedigree. The pedigree
is three character vectors, with unknown parents written as `NA` or as
whatever `unknownParent` says, and it is extended backwards over the
parent names that carry a relationship. Founders are
either sampled at random, with `matchID = FALSE`, or matched by name to
genotyped individuals, with `matchID = TRUE`. Given a map population it
builds and discards its own `SimParam` and returns a `NamedMapPop`, which
makes the import, drop, simulate sequence the natural way to start a
simulation from a real breeding population.
