---
title: "Parallelization in AlphaSimR"
author: "Gregor Gorjanc and Chris Gaynor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallelization in AlphaSimR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

`AlphaSimR` leverages `OpenMP`-type parallelization
to speed-up execution of some underlying `C++` code called by `R` code.
This speed-up can be substantial for large-scale simulations.
Depending on the version and source of your installed `R` and `AlphaSimR`,
`OpenMP` might or might not be supported out of the box.
In our experience,
it is usually available automatically on Windows and Linux.
macOS users most often need to take extra steps.
We show below how to check if you can use multiple threads and
how these get used in `AlphaSimR`.
We also point to more information on installation/building to enable `OpenMP`.

## Check whether OpenMP is available

The simplest check is via `getNumThreads()`.

```{r}
library(AlphaSimR)
getNumThreads()
```

If `OpenMP` is not available (or when you only have one available thread),
`getNumThreads()` returns `1` and all code will run single-threaded.
When `getNumThreads()` returns more than `1`,
some code will run in parallel.

## Controlling the number of threads

When `OpenMP` is enabled,
`AlphaSimR` uses all the available threads by default.
You can control the number of threads used
via the `nThreads` argument
globally in the `SP` object or
individually in some functions.

```{r}
founderPop = quickHaplo(nInd = 4, nChr = 2, segSites = 5)
SP = SimParam$new(founderPop)
SP$nThreads
```

Assuming that you have more than one thread,
but want to use just one globally,
you can specify that in the `SP` object:

```{r}
pop = newPop(founderPop, simParam = SP)
SP$nThreads = 1L
pullSegSiteGeno(pop)
# Note that pullSegSiteGeno() uses internally SP$nThreads,
# when the nThreads argument is not specified.
```

As indicated above,
we can specify the number of threads in some functions:

```{r}
pullSegSiteGeno(pop, nThreads = 1L)
```

## A demonstration

There are several `AlphaSimR` functions that leverage parallelization,
particularly when working across chromosomes or
with linear algebra operations.
`runMacs()` is one such example.
The code below reports time for simulating genomes with 10 chromosomes for 50 individuals
in serial mode and in parallel mode.
You can evaluate the speed-up due to parallelization by
changing the number of chromosomes,
and also increasing the size of simulation by increasing the number of individuals.

```{r}
system.time(runMacs(nInd = 50, nChr = 10, nThreads = 1L))
system.time(runMacs(nInd = 50, nChr = 10, nThreads = getNumThreads()))
```

## How to get OpenMP support

This topic depends on your platform and `R` toolchain,
so we keep the guidance short and point to the relevant docs:

- Windows and Linux: in our experience, `OpenMP` is usually enabled automatically.
  Install `AlphaSimR` and check `AlphaSimR::getNumThreads()`.

- macOS: if `OpenMP` is not already available,
  follow the instructions at <https://mac.r-project.org/openmp/>,
  then reinstall `AlphaSimR` from the `CRAN` or `GitHub` source, and
  then check `AlphaSimR::getNumThreads()`.

## A note on `BLAS`

For linear algebra-heavy work (such as fitting genomic prediction models),
a multithreaded `BLAS` can also help.
This is separate to `OpenMP` support.
In our experience, this is even more platform- and `R`-build-specific than `OpenMP`,
so it is best treated as an advanced local optimisation.

- Windows: We have had reasonable results with `OpenBLAS`:
  <https://github.com/OpenMathLib/OpenBLAS>.

- Linux: `BLAS` setup depends on how `R` was installed.
  You may be able to replace the shared `BLAS` library,
  or compile `R` against a specific `BLAS`.
  See the `R Installation and Administration` manual:
  <https://cran.r-project.org/doc/manuals/r-release/R-admin.html#BLAS>.

- macOS: see the `R for macOS FAQ`:
  <https://cran.r-project.org/bin/macosx/RMacOSX-FAQ.html#Which-BLAS-is-used-and-how-can-it-be-changed_003f>.
  On recent `R` versions, simple `BLAS` library swapping may not work cleanly,
  so follow the guidance for your specific `R` build and verify that base `R`
  still starts correctly after any change.
