---
title: "Quantitative Genetics in AlphaSimR"
author: "Gregor Gorjanc and Chris Gaynor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: AlphaSimR.bib
vignette: >
  %\VignetteIndexEntry{Quantitative Genetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(AlphaSimR)
```

This article describes how AlphaSimR implements quantitative genetics
theory. The companion article on traits describes the *functional* model:
how a genotype is turned into a genetic value through additive, dominance
and epistatic effects. This article describes the *statistical* model:
how those genetic values are decomposed into breeding values, dominance
deviations and epistatic deviations, and how the variance of a population
is partitioned into components. You should be familiar with the traits
article before reading this one.

The distinction between the two models is the central idea of the field
and is worth stating plainly at the outset. The functional model is a
property of the biology: it says what genetic value a given genotype has.
The statistical model is a property of a *population*: it says how much
of the variation among the individuals in that population can be
attributed to effects that are transmitted from parent to offspring. The
same functional model, evaluated in two populations with different allele
frequencies, yields different breeding values and different variance
components. Everything below follows from taking that seriously.

All of the quantities described here are computed by `genParam`, which
returns them for every trait in the simulation.

## The genotypic model

AlphaSimR builds the genetic value of an individual from biallelic loci
using scaled genotype dosages. For a locus with raw dosage $x$ (the
number of copies of the "1" allele) in an organism of ploidy $P$, the
scaled additive and dominance dosages are

$$x_A = \left( x - \frac{P}{2} \right)\left( \frac{2}{P} \right),
\qquad
x_D = x \left( P - x \right) \left( \frac{2}{P} \right)^{2}$$

and the genetic value of an individual is

$$G = \mu_0 + \sum_i a_i x_{A_i} + \sum_i d_i x_{D_i} +
\sum_{(i,j)} e_{ij} x_{A_i} x_{A_j}$$

where $\mu_0$ is the trait intercept, $a$ and $d$ are additive and
dominance effects, and $e$ is an additive-by-additive epistatic effect.
The epistatic sum runs over *pairs* of QTL. AlphaSimR assigns every QTL
to exactly one pair, so the pairs are disjoint and the number of pairs is
half the number of QTL. This matters for everything that follows, because
it means the two-locus calculations described below can be carried out
one pair at a time and summed.

Note that the only epistasis in the model is additive-by-additive.
There are no additive-by-dominance or dominance-by-dominance terms.

## Breeding values

The breeding value of an individual is the part of its genetic value that
it transmits to its offspring. Formally it is the least-squares
projection of the genetic value onto allele dosage, and the coefficient
of that projection is the average effect of an allele substitution,
usually written $\alpha$ [@FalconerMackay].

Consider a single locus in a population where genotype dosage $x$ occurs
with frequency $f(x)$. Let $\bar{x} = \sum_x f(x)\, x$ be the mean dosage
and $G(x)$ the genetic value of dosage $x$. The average effect of an
allele substitution is the regression coefficient of genetic value on
dosage,

$$\alpha = \frac{\sum_x f(x)\,\big(G(x) - \bar{G}\big)\big(x -
\bar{x}\big)}{\sum_x f(x)\,\big(x - \bar{x}\big)^{2}}$$

and the breeding value of an individual carrying dosage $x$ is

$$A(x) = \big(x - \bar{x}\big)\, \alpha .$$

An individual's breeding value for the trait is the sum of $A(x)$ over
all QTL. Breeding values are centred on the population mean by
construction, since $\sum_x f(x)(x - \bar{x}) = 0$.

Two features of this definition deserve emphasis. First, $\alpha$ is not
a property of the locus alone. It depends on $f(x)$, and therefore on
allele frequency and on the mating system. A locus with a large dominance
effect can have $\alpha = 0$ at one allele frequency and a substantial
$\alpha$ at another. Second, when there is no dominance and no epistasis,
$G(x)$ is linear in $x$, the regression is exact, and $\alpha$ reduces to
the additive effect rescaled by the dosage scaling. Breeding values then
account for all of the genetic variation.

AlphaSimR computes two versions of $\alpha$ for every locus. The first
uses the genotype frequencies actually observed in the population and is
returned as `alpha`. The second uses the genotype frequencies expected
under Hardy-Weinberg equilibrium at the observed allele frequency, and is
returned as `alpha_HW`. Writing $p = \bar{x}/P$ for the allele frequency,
the Hardy-Weinberg genotype frequencies for ploidy $P$ are

$$f_{HW}(x) = \binom{P}{x} p^{x} (1-p)^{P-x}$$

which is the familiar binomial expansion, and which for $P = 2$ gives the
usual $p^2$, $2pq$, $q^2$. The two versions of $\alpha$ differ whenever
the population departs from Hardy-Weinberg proportions, which is to say
whenever there is inbreeding or any other form of non-random mating. The
difference between them is what AlphaSimR later reports as the
non-random-mating component of variance.

```{r}
founderPop = quickHaplo(nInd = 100, nChr = 2, segSites = 20)
SP = SimParam$new(founderPop)
SP$addTraitAD(nQtlPerChr = 10, meanDD = 0.5)
pop = newPop(founderPop, simParam = SP)

ans = genParam(pop, simParam = SP)

head(ans$alpha[[1]])    # Observed frequencies
head(ans$alpha_HW[[1]]) # Hardy-Weinberg frequencies
```

## Dominance deviations

Once the breeding value has absorbed everything that can be explained by
a linear regression on dosage, whatever is left over at that locus is the
dominance deviation. It is defined as a lack of fit,

$$D(x) = G(x) - A(x) - \bar{G}$$

so that $G(x) = \bar{G} + A(x) + D(x)$ exactly. Because $A(x)$ is a
least-squares projection, the dominance deviations are uncorrelated with
the breeding values within a locus, and they have mean zero over the
population.

The name is slightly misleading in the way that all of Fisher's
terminology is slightly misleading. The dominance deviation is not "the
dominance effect". It is the residual from the additive regression, and
it exists only because the functional dominance effect $d$ makes $G(x)$
non-linear in $x$. If $d = 0$ then $D(x) = 0$ for every genotype
regardless of allele frequency. Conversely, a locus with a large $d$
contributes to the additive variance as well as the dominance variance,
and at extreme allele frequencies most of its contribution is additive.

As with $\alpha$, AlphaSimR computes dominance deviations twice, once
against the observed genotype frequencies and once against
Hardy-Weinberg expectations.

## Epistatic deviations

Epistatic deviations follow the same lack-of-fit logic, applied to a pair
of loci rather than to a single locus. The calculation proceeds in two
stages.

The first stage marginalises. For the pair $(i,j)$, the marginal genetic
value of dosage $x_i$ at the first locus is its genetic value averaged
over the distribution of dosages at the second locus,

$$G_i(x_i) = \sum_{x_j} f(x_j) \, G(x_i, x_j)$$

and symmetrically for the second locus. Because the epistatic term is
$e\, x_{A_i} x_{A_j}$, the average over the second locus contributes
$e\, x_{A_i} \bar{x}_{A_j}$, which is linear in $x_{A_i}$. The epistatic
effect therefore shifts the marginal genetic value at the first locus in
a way that depends on the allele frequency at the second locus. This is
the mechanism by which functional epistasis generates statistical
additive variance.

The average effect $\alpha_i$ and the breeding value $A_i$ are then
obtained from the marginal genetic values exactly as in the single-locus
case, and the dominance deviation $D_i$ is the residual from that
regression. Note that because the marginalised epistatic contribution is
linear in $x_{A_i}$, it is absorbed entirely by the additive regression
and leaves the dominance deviation untouched. Epistasis in this model
generates additive and epistatic variance, but no dominance variance.

The second stage assembles the pair. The two-locus breeding value and
dominance deviation are the sums of the single-locus quantities,

$$A(x_i, x_j) = A_i(x_i) + A_j(x_j), \qquad
D(x_i, x_j) = D_i(x_i) + D_j(x_j)$$

and the epistatic deviation is what remains,

$$AA(x_i, x_j) = G(x_i, x_j) - A(x_i, x_j) - D(x_i, x_j) - \bar{G}.$$

Summing over pairs gives the individual's epistatic deviation for the
trait. As with the other components it has mean zero over the population,
so that for every individual

$$G = \mu + A + D + AA$$

holds exactly. This identity is worth verifying, because it is the
clearest statement of what the decomposition is doing.

```{r}
founderPop = quickHaplo(nInd = 100, nChr = 2, segSites = 20)
SP = SimParam$new(founderPop)
SP$addTraitADE(nQtlPerChr = 10, meanDD = 0.5, relAA = 0.2)
pop = newPop(founderPop, simParam = SP)

ans = genParam(pop, simParam = SP)

# G = mu + A + D + AA, for every individual
resid = ans$gv[,1] - (ans$mu[1] + ans$bv[,1] + ans$dd[,1] + ans$aa[,1])
range(resid)
```

## Partitioning the variance

Having decomposed each individual's genetic value, the variance can be
partitioned. AlphaSimR does this against three different reference
populations, and keeping them straight is the key to reading the output
of `genParam`.

The **observed population** is the set of individuals actually in `pop`,
with their actual joint genotypes. Variances computed here are the
*genetic* variances: `varA`, `varD`, `varAA` and `varG`. They are simply
the variances of the breeding values, dominance deviations, epistatic
deviations and genetic values across individuals, and they include every
correlation between loci that exists in the population.

The **no-linkage-disequilibrium population** has the same genotype
frequencies at each locus as the observed population, but loci are
treated as independent. Variances computed here are sums of single-locus
(or single-pair) contributions with no covariance between loci.

The **idealised population** has the same *allele* frequencies as the
observed population, but genotype frequencies at Hardy-Weinberg
proportions and loci in linkage equilibrium. Variances computed here are
the *genic* variances: `genicVarA`, `genicVarD`, `genicVarAA` and
`genicVarG`. The genic variance is the variance the population would have
if it were randomly mated and its loci were unlinked, holding allele
frequencies fixed.

The genic variances are computed as exact expectations over the frequency
distribution rather than as sample variances over individuals. For the
additive component,

$$\sigma^{2}_{A,\text{genic}} = \sum_i \sum_x f_{HW}(x)\, A_i(x)^{2}$$

and similarly for the dominance and epistatic components. Because the
deviations have mean zero, these sums of squares are variances.

## The Bulmer partitioning

The genetic variance and the genic variance are not the same, and the
difference between them is informative. Selection and non-random mating
change the genetic variance of a population long before they change its
allele frequencies, because they generate correlations between loci and
departures from Hardy-Weinberg proportions. This is the Bulmer effect,
described by @bulmer_1971, and it is the reason a breeding programme
loses additive variance faster than allele-frequency change alone would
suggest.

`genParam` reports the difference as two separate covariance terms, so
that for each component

$$\sigma^{2}_{A} = \sigma^{2}_{A,\text{genic}} +
\text{cov}_{A,HW} + \text{cov}_{A,L}$$

with the analogous identity for the dominance and epistatic components.
The three terms are:

`genicVarA`, the variance of an idealised population at the same allele
frequencies, randomly mated and in linkage equilibrium. This changes only
when allele frequencies change, through selection, drift or mutation.

`covA_HW`, the covariance arising from departures from Hardy-Weinberg
proportions. It is computed as the difference between the genic variance
evaluated at the observed genotype frequencies and the genic variance
evaluated at Hardy-Weinberg genotype frequencies. Inbreeding makes this
term positive for the additive component: an inbred population has more
additive variance than a randomly mated one at the same allele
frequencies, because homozygotes are over-represented.

`covA_L`, the covariance between loci arising from linkage
disequilibrium. It is obtained by subtraction, as the genetic variance
minus the genic variance minus the Hardy-Weinberg term. Directional
selection generates negative linkage disequilibrium among loci affecting
the trait, so this term is typically negative in a selected population,
and it is the term that carries the Bulmer effect.

Because the total genetic value is a sum of three components, the total
genetic variance also includes covariances *between* components. These
are reported separately as `covAD_L`, `covAAA_L` and `covDAA_L`, the
covariances between breeding values and dominance deviations, between
breeding values and epistatic deviations, and between dominance and
epistatic deviations. All three are zero in the idealised population and
non-zero in general. The complete identity is

$$\sigma^{2}_{G} = \sigma^{2}_{A} + \sigma^{2}_{D} + \sigma^{2}_{AA}
+ 2\,\text{cov}_{AD,L} + 2\,\text{cov}_{A\cdot AA,L}
+ 2\,\text{cov}_{D\cdot AA,L}$$

which follows directly from $G = \mu + A + D + AA$ and holds numerically.

```{r}
# The full variance identity
varG = ans$varG[1,1]
parts = ans$varA[1,1] + ans$varD[1,1] + ans$varAA[1,1] +
  2*(ans$covAD_L[1] + ans$covAAA_L[1] + ans$covDAA_L[1])

c(varG = varG, parts = parts)
```

The Bulmer partitioning is easiest to see by selecting and then looking
at what changed. In the example below the genic additive variance is
almost unchanged after one generation of strong selection, because allele
frequencies have barely moved, while the additive genetic variance has
dropped because selection has generated negative linkage disequilibrium.

```{r}
set.seed(1234)

founderPop = quickHaplo(nInd = 1000, nChr = 10, segSites = 20)
SP = SimParam$new(founderPop)
SP$addTraitA(nQtlPerChr = 10)
SP$setVarE(h2 = 0.5)
pop = newPop(founderPop, simParam = SP)

sel = selectInd(pop, nInd = 100, simParam = SP)
prog = randCross(sel, nCrosses = 1000, simParam = SP)

before = genParam(pop, simParam = SP)
after = genParam(prog, simParam = SP)

round(c(varA_before = before$varA[1,1],
        varA_after = after$varA[1,1],
        genic_before = before$genicVarA[1],
        genic_after = after$genicVarA[1],
        covL_before = before$covA_L[1],
        covL_after = after$covA_L[1]), 3)
```

## Polyploids and digenic dominance

The dominance model requires special comment for autopolyploids, because
a polyploid locus has more genotypes than a fully specified dominance
model can be fitted with from a single parameter.

A diploid locus has one heterozygous genotype, and a single dominance
effect $d$ describes it completely. An autotetraploid locus has three
heterozygous genotypes, with dosages 1, 2 and 3. Describing all three
requires three parameters, conventionally decomposed into digenic,
trigenic and quadrigenic dominance, corresponding to interactions among
two, three and four alleles at the locus. @Gallais gives the full
treatment.

AlphaSimR uses one parameter. The scaled dominance dosage
$x_D = x(P-x)(2/P)^2$ is proportional to $x(P-x)$, which is the number of
heterozygous pairs of alleles among the $P$ alleles at the locus: each of
the $x$ copies of the "1" allele pairs with each of the $P-x$ copies of
the "0" allele. The dominance contribution is therefore proportional to
the count of heterozygous *pairs*, which is precisely a digenic dominance
model. Interactions involving three or four alleles at once are not
represented.

| Diploid dosage | Tetraploid dosage | $x_A$ | $x_D$ |
|---------------:|------------------:|------:|------:|
| 0              | 0                 | $-1$    | $0$    |
|                | 1                 | $-1/2$  | $3/4$  |
| 1              | 2                 | $0$     | $1$    |
|                | 3                 | $1/2$   | $3/4$  |
| 2              | 4                 | $1$     | $0$    |

The consequence for the statistical decomposition is that the dominance
deviations of a polyploid span only the digenic subspace. The
decomposition itself is still exact: $G = \mu + A + D + AA$ holds for
polyploids as it does for diploids, and the dominance deviations are
still the least-squares residual from the additive regression. What is
restricted is the functional model being decomposed. A real
autotetraploid whose loci show trigenic or quadrigenic dominance cannot
be represented, and the dominance variance AlphaSimR reports for a
polyploid is the digenic dominance variance only.

There is a second polyploid-specific caveat, this one concerning the
Hardy-Weinberg reference. The binomial genotype frequencies used to
compute the genic variances are the correct polysomic equilibrium only
under random chromosome segregation without double reduction. Quadrivalent
pairing, described in the article on genetic recombination, produces
double reduction and therefore an excess of homozygotes relative to the
binomial expectation. AlphaSimR does not model this in its
Hardy-Weinberg reference, so in a simulation with `SP$quadProb > 0` the
resulting excess homozygosity appears in the `covD_HW` and `covA_HW`
terms as though it were non-random mating. This is arguably the right
place for it, since the effect on the variance is the same, but it means
those terms should not be read as a pure measure of the mating system in
an autopolyploid simulation.

```{r}
founderPop = quickHaplo(nInd = 100, nChr = 2, segSites = 20, ploidy = 4L)
SP = SimParam$new(founderPop)
SP$addTraitAD(nQtlPerChr = 10, meanDD = 0.5)
pop = newPop(founderPop, simParam = SP)

ans4 = genParam(pop, simParam = SP)
round(c(varA = ans4$varA[1,1],
        varD = ans4$varD[1,1],
        genicVarA = ans4$genicVarA[1],
        genicVarD = ans4$genicVarD[1]), 3)
```

## Relationship to the classical theory

The decomposition described here is the least-squares partition
introduced by @fisher_1919 and extended to epistasis by @cockerham_1954
and @kempthorne_1954. The additive-by-additive contrast used by
AlphaSimR is the product of the two single-locus additive contrasts,
which is the standard construction and the one that makes the components
orthogonal in an idealised population. In the terminology of the natural
and orthogonal interactions framework of @alvarezcastro_2007,
AlphaSimR computes the statistical decomposition against the observed
genotype frequencies, which is what makes `alpha` and `alpha_HW`
differ.

Orthogonality of the components holds exactly only in the idealised
population, with Hardy-Weinberg proportions and linkage equilibrium. In
any real or simulated population the components are correlated, and
AlphaSimR does not pretend otherwise: the correlations are reported
explicitly as the `cov*_L` terms rather than being assumed away. This is
why the sum of `varA`, `varD` and `varAA` does not equal `varG` unless
the cross-component covariances are included.

One further point of interpretation, made forcefully by @hill_2008, is
worth keeping in mind when setting up a simulation with epistasis. A
functional model with substantial additive-by-additive epistasis will
usually produce a population in which most of the genetic variance is
statistically additive, particularly when allele frequencies are away
from one half. The `relAA` argument specifies the ratio of
epistatic to additive variance expected in a population at allele
frequency 0.5 in linkage equilibrium; observed ratios in a simulated
population will typically be smaller, sometimes much smaller. This is not
an artefact of the implementation but a property of the decomposition
itself.

## References

